---
title: 'Looping'
author: "Naomi Tague"
date: last-modified
format:
  revealjs:
    theme: solarized
    resources: ["images"]
editor: visual
execute:
  echo: true
  warning: true
  message: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE)

library(tidyverse)
library(purrr)
library(ggpubr)
```

## [Looping]{style="color:orange"}

A programming *building block*

Used to repeat instructions. In environmental science, often used for calculations that involve:

-   space (loop over location)
-   time (loop through time)
-   time and space

Questions?

## [Steps looping]{style="color:orange"}

1.  design a data structure to store results: sometimes this is automatic but not always
2.  read in (or generate test data) the data
3.  find ways to repeat things efficiently

## [Looping 'behind' the scene]{style="color:orange"}

-   column manipulation with data frames
-   *dplyr* and *purrr* family of functions in R

## [Nesting Loops]{style="color:orange"}

Loops can be "nested" with one loop inside the other

Examples:

-   doing something to each row AND each column
-   doing something to each location AND each time step
-   doing something across categories AND through time

## Syntax in R {.scrollable}

```         
FOR (i in 1:nrow) {

  FOR (j in 1:ncol) {

    ...

  }
}
```

For each single step of the outer loop, the inner loop runs through **all** its iterations

So the loop above processes all columns for the first row, then moves to the next row

As before you don't need to use *i* and *j* - these are just common variable names for counters but you can pick more meaningful ones

## Simple Example of Nested Loops {.scrollable}

```{r}
# simple example of nested loops - compute
# the sum of all the elements in a matrix keeping track of something at each location
# create some sample data
ntime = 5
nlocation = 3
# matrix function in R lets you initialize with data
# here we are creating a matrix with nlocation rows and nobs columns, filled with integers between 1 and 10
creatureobs = matrix(sample(1:10, ntime*nlocation, replace=TRUE), nrow=nlocation)
colnames(creatureobs) = c("time1", "time2", "time3", "time4", "time5")
rownames(creatureobs) = c("east", "west", "north")
# check to see that it is a matrix with rows and columns
str(creatureobs)
creatureobs

# nested for loop: use t as counter for time; l as counter for location
# initialization
# outer loop for location
totalsum = 0
for (l in 1:nlocation) {
  # inner loop for observations
  for (t in 1:ntime) {
     totalsum = totalsum + creatureobs[l,t]
  }
}
totalsum

## to see if work print out at each location and time step
totalsum = 0
for (l in 1:nlocation) {
  # inner loop for observations
  for (t in 1:ntime) {
     totalsum = totalsum + creatureobs[l,t]
     # use sprintf to print out the location, time, and sum so far, cat just prints out the string
     cat(sprintf("\nfor counter l=%d t=%d, location: %s, time: %s, sum so far: %s", l, t, rownames(creatureobs)[l], colnames(creatureobs)[t], totalsum))
  }
}

totalsum
# R can do these for loops internally when you do
sum(creatureobs)

```

## [Slightly more complex example]{style="color:orange"} {.scrollable}

Imagine we have running speeds of a species for different points in a transect

Observations are given to you in a matrix where rows are locations in the transect and columns are observations of speeds at each location for different individuals

1.  Find number of times speeds were above the hunting threshold anywhere (easy)

-   just count all values in the matrix above threshold (e.g. `sum(matrix >= threshold)`)
-   could do without a loop

2.  Find how often neighboring locations had speeds above the hunting threshold at the same time (harder)

-   need to loop through each location and observation and check if neighboring location also had a speed above the threshold at that time

-   loops make this easy because you can go back and forth in time with your counter!

## Start with #1 {.scrollable}

*find number of times speeds were above the hunting threshold anywhere (easy)*

try with a partner

```{r}

# create some sample data
nobs = 4
nlocation = 5
# matrix function in R lets you initialize with data
creatureobs = matrix(runif(nobs*nlocation, min=0, max=100), nrow=nlocation)
# check to see that it is a matrix with rows and columns
str(creatureobs)
creatureobs

# nested for loop: use o as counter for observation; l as counter for location
# outer loop for location
hunting_speed = 60
num_hunting_speeds=0
for (l in 1:nlocation) {
  # inner loop for observations
  for (o in 1:nobs) {
     num_hunting_speeds = num_hunting_speeds + ifelse(creatureobs[l,o] >= hunting_speed, 1, 0)
  }
}

num_hunting_speeds
# as percent
num_hunting_speeds/(nobs*nlocation)*100

```

## Now let's try the harder task {.scrollable}

```{r}
# nested for loop: use o as counter for observation; l as counter for location
# outer loop for location
num_hunting_speeds_neighbors=0

# notice that we only go to the second last location
for (l in 1:(nlocation-1)) {
  # inner loop for observations
  for (o in 1:nobs) {
      if (creatureobs[l,o] >= hunting_speed & creatureobs[l+1,o] > hunting_speed) {
         checkneigh = 1
      } else {
         checkneigh = 0
      }
     num_hunting_speeds_neighbors = num_hunting_speeds_neighbors + checkneigh
  }
}
num_hunting_speeds_neighbors
# as percent of all observations
num_hunting_speeds_neighbors/((nobs*(nlocation-1)))*100

```

## [A bit more on nesting loops]{style="color:orange"} {.scrollable}

We could even have multiple levels of nesting — think about doing something to a 3-dimensional matrix of time and space (x, y)

```         
FOR (x in 1:nrow) {

  FOR (y in 1:ncol) {

    FOR (t in 1:ntime) {

      ...

    }
  }
}
```

To illustrate

```{r}

# a 3-dimensional array: space (x=20, y=30) and time (10 steps)
ocean = array(dim=c(20,30,10))
str(ocean)

# fill with simulated temperature data using nested loops
for (x in 1:nrow(ocean)) {
  for (y in 1:ncol(ocean)) {
    for (t in 1:dim(ocean)[3]) {
      ocean[x,y,t] = rnorm(mean=16+t, sd=2, n=1)
      # imagine running a model to estimate ocean temperature
      # then feeding it into a model that uses that temperature to impact corals
    }
  }
}

```

## [Different types of loops]{style="color:orange"}

-   *for* - works with a counter to decide how many times to repeat

-   *while* - repeats until a condition is met

## [While loop]{style="color:orange"}

Goes through the loop until a condition is met

-   useful for when you don't know how many times you need to repeat something but you want to keep going until a condition is met
-   often for simulations where you want to keep going until you reach a threshold or a certain outcome

## Format for *while*

```         
while (condition) {

  ...

}
```

## Example of **while** loop {.scrollable}

Imagine trying to estimate how long it will take for a water quality threshold to be exceeded when there is a leak of pollution into a lake;

-   you have an estimate of how much pollution is added each day; (**pol_in**)

-   You also know that the pollution is removed at a certain rate each day; (**pol_out**) but removal efficacy depend on the air temperature that day

    -   removal is $pol_{out} = pol_{out,base} * (1 + 0.01 * air_{temp})$

-   You know air temperature is normally distributed mean 20 degrees Celsius and standard deviation of 5 degrees Celsius

you want to know how long it will take for the pollution to exceed a threshold

**NOTE:** I'm also going to add a check to stop if too many days pass! (we will talk about why)

```{r, whileexample, error=FALSE}
# initialize parameters
pol_in = 10 # pollution added each day
pol_out_base = 5 # base pollution removed each day
threshold = 100 # pollution threshold

# initialize pollution level and day counter
pollution_level = 0
day = 0

maxday =365
# while loop to simulate pollution levels over time
# we use && for AND because we want to check both conditions - that pollution level is below threshold and that we haven't exceeded max days

while ((pollution_level < threshold) && (day < maxday)) {
  # update pollution level
  air_temp = rnorm(mean=20, sd=5, n=1) #
  pol_out = pol_out_base * (1 + 0.01*air_temp) # calculate pollution removed based on air temp
  pollution_level = pollution_level + pol_in - pol_out # update pollution level
  day = day + 1 # increment day counter
}
cat(sprintf("It takes %d days for pollution to exceed the threshold of %d, final level: %.1f", day, threshold, pollution_level))
```

## [Why do we need a check for max days?]{style="color:orange"}

What if pollutant level never exceeds threshold?

Infinite Loop!

you can often use 'ctrl C' to break out of an infinite loop

it is better to have a check in your code to stop if you exceed a certain number of iterations

## Reminder for first make up recorded lecture
