---
title: 'Functions'
format:
  revealjs:
    theme: solarized
    resources: ["images"]
editor: visual
execute:
  echo: true
  warning: true
  message: true
---

```{r, setup, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
```

## \# [Lets practice:: Automobile Power Generation]{style="color:orange"}

One of the equations used to compute automobile fuel efficiency is given below. This is the power required to keep a car moving at a given speed, and is as follows:

$Pb = crolling * m *g *V + \frac{1}{2} * A*p_{air}*c_{drag}*V^3$

where crolling and cdrag are rolling and aerodynamic resistive coefficients, typical values are 0.015 and 0.3, respectively.

- V: is vehicle speed (assuming no headwind) in m/s (or mps) m: is vehicle mass in kg

- A is surface area of car $(m^2)$

- *g*: is acceleration due to gravity $(9.8 m/s^2 )$

- p_air = density of air $(1.2kg/m^3 )$

- Pb is power in Watts

What would be the function's contract - input/parameters? what will it do?

# [Lets practice:: Automobile Power Generation]{style="color:orange"}

1.  Write a function to compute power, given a truck of m=31752 kg (parameters for a heavy truck) for a range of different highway speeds, the surface area of this truck is 16 $m^2$

- 30 m/s
- 80 km/hr
- 120 km/hr
- 40 km/hr
- 5 km/hr

2.  Plot power as a function of speed

3.  Recompute for a lighter vehicle (15,000 kg) Add a second line to your graph for lighter vehicle

- Keep your function in its own file, store it in a subdirectory called *R* and then create a **separate** Rmarkdown (or Quarto) document to use the function and store the Rmarkdown in a different directory

# **Good Programming Practices**

Name variables such that they are easy to understand

Describe the contract at the top of the function (following the format for the *power_gen_orig.R* example)

Also include the comments in the Quarto

# What we've learned

- how to write a function (and add error checking)
- how to generate data
- how to repeat in code (different types of looping)

# A more complex example of looping in R

Lets **scale** up from a single car to a group of cars on a highway and use our **autopower** function to estimate a distribution of power

What might vary?

# Our highway

3 car types

Imagine with have 3 different car types - and we know how often each occurs:

- car A mass 20000 kg, area 22 m^2^

- car B mass 65000 kg, area 30 m^2^

- car C mass 38000 kg area 22 m^2^

Mean highway speed is 100 km/hr

```{r sampling2, eval=TRUE, echo=TRUE}
library(here)
library(tidyverse)
source(here("course-materials/R/autopower.R"))
# generate a structure to store info on our possible cars
possible_cars = data.frame(name = c("A","B","C"),mass=c(10000,65000,38000), area = c(22,30,22))

# first look at how results vary for mean speed say 100km/hr
# do conversion
speed_base = 100 * 0.28

# because I have one mass and area for each car and only 1 speed
# I can estimate power for each car type
# add to the data structure
possible_cars$power = autopower(V=speed_base, A = possible_cars$area, m=possible_cars$mass)

head(possible_cars)

# show results                         
ggplot(possible_cars, aes(x=mass, y=power, fill=as.factor(area)))+geom_col()+labs(y="Power W", x="Mass (kg)", fill="Area (m2)")
  

```

# Building a highway

What could be the total power consumed if there are 200 cars on this highway each hour, they are travelling at a range of speeds - mean is 100km/hr and speeds tend to be log-normally distributed)

How would the total power consumed vary by car So if all cars are car A; OR all cars are car B OR all cars are car C

We will use *sample* here to generate speeds for our 100 cars and look at different ways to repeat power calculation for different cars

- repeating by hand
- *for* for repetition - a more standard way of looping available in many langugaes
- *pmap* for repetition - a efficient way of looping in R

# First lets do it 'by hand'

```{r byhand}

# what is I want to estimate average power use given  each car

possible_cars

# use sample to generate a distribution of speeds

# assume a log normal distribution of speeds with mean 100km/hr, and standard deviation that is 10% of the mean

# recall our function needs speed in m/s not km/hr so we will also do a conversion
# lets get a sample of a 200 speeds- we could also do this by actually measuring speeds

nsample = 200
mean_speed = log(100*0.277)

speeds = rlnorm(mean=mean_speed, sd=0.1*mean_speed, nsample)
summary(speeds)

plot(density(speeds), ylab="Distribution of Speeds in (m/s)")


# how do we run each car for all speeds 

# first lets to it by hand for the first car - the first row in our possible cars matrix
possible_cars[1,]

# we could do it by hand
powerA = autopower(V=speeds, A = possible_cars$area[1], m=possible_cars$mass[1])
# lets look at what we get
summary(powerA)

# next car (row 2)
powerB = autopower(V=speeds, A = possible_cars$area[2], m=possible_cars$mass[2])
# lets look at what we get
summary(powerB)

# next car (row 3)
powerC = autopower(V=speeds, A = possible_cars$area[3], m=possible_cars$mass[3])
# lets look at what we get
summary(powerC)


# we could put this together
powerall1 = cbind.data.frame(powerA, powerB, powerC)
colnames(powerall1)=possible_cars$name


# for plotting sometimes its useful to turn columns in to rows - we can use an R function
# called pivot_longer (part of the tidyverse package) to do this
# compare powerall1 and powerallr1 to see what pivot_longer does
powerallr1 = powerall1 %>% pivot_longer(cols=everything(), names_to="car", values_to="power")
head(powerallr1)

# quick visualization
# lets save it so that we can compare
method1_plot = ggplot(powerallr1, aes(car,power, fill=car))+geom_boxplot()+ggtitle("By Hand")
method1_plot



```

# Second- classic looping *for*

Doing this by hand would be hard if we had many different cars - can we automate?

YES

For loop

Lets do this one more time using a *for* loop

```{r withforloop}
library(ggpubr)
# the first part, generating speeds is the same
# what is I want to estimate average power use given  each car

possible_cars


# how do we run each car for all speeds 
# we use a for loop to cycle through
# we need to create a data frame to store results - as above
# one column for each car and one row for each speed

powerall2 = as.data.frame(matrix(nrow=length(speeds), ncol=nrow(possible_cars)))
# because we don't initialize it - values are NA
head(powerall2)

# how many cars area there
nrow(possible_cars)
ncol(powerall2)
# for loops use an index - in this case "i" but you could use anything - it repeats
# anything between the {} for each values of i between 1 and nrow(possible_car) (which is 3 in our case)

# index in to a matrix (like powerall3) is by row and column powerall3[2,5] is 2nd row and 5th column
for (i in 1:ncol(powerall2)) {
  powerall2[,i] = autopower(A=possible_cars$area[i], m=possible_cars$mass[i], V=speeds)
}

# now it looks like above
head(powerall2)
# we can add column names
colnames(powerall2)=possible_cars$name

# plotting is the same as above

# for plotting sometimes its useful to turn columns in to rows
powerallr2 = powerall2 %>% pivot_longer(cols=everything(), names_to="car", values_to="power")
head(powerallr2)

# quick visualization
method2_plot = ggplot(powerallr2, aes(car,power, fill=car))+geom_boxplot()+ggtitle("For Loop")

ggarrange(method1_plot, method2_plot, nrow=2)

powerall1 %>% map(mean)
powerall2 %>% map(mean)

```

# Third using R built in tools

*pmap* works quickly but it is unique to R Other programming language (and R) use what are called loops - where repetition is more explicit

*pmap* is available in the *purr* library

*mapply* is another R option

*pmap* - runs a function for a list of parameters - using other parameters for each iteration

```{r withpmap}



# how do we run each car for all speeds 
# pmap runs a function for each value in a list of parameters, with other parameters set for each iteration


powerall3 = pmap(list(A = possible_cars$area, m=possible_cars$mass), autopower, V=speeds)

# lets turn to a data frame for easier graphing
# we can add column names
powerall3 = as.data.frame(powerall2, col.names=possible_cars$name)

head(powerall3)


# for plotting sometimes its useful to turn columns in to rows
powerallr3 = powerall3 %>% pivot_longer(cols=everything(), names_to="car", values_to="power")
head(powerallr2)

# quick visualization

method3_plot = ggplot(powerallr3, aes(car,power, fill=car))+geom_boxplot()+ggtitle("pmap")
method3_plot

# put plots side by side
# to confirm that they look similar
ggarrange(method1_plot, method2_plot, method3_plot, nrow=3)

# compare values
head(powerallr2)
head(powerallr1)
head(powerallr3)


```
