---
title: 'Functions'
format:
  revealjs:
    theme: solarized
    resources: ["images"]
editor: visual
execute:
  echo: true
  warning: true
  message: true
---

```{r, setup, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
```

## [Loops and Functions]{style="color:orange"}

Calculate NPV for

-   a range of different interest rates and
-   a range of damages
-   that may be incurred 10 years in the future

**Steps**

-   define inputs (interest rates, damages)
-   define output (NPV)
-   write the function
-   create a data structure to store results where we vary both interest rates and damages
-   use nested for loops to fill in the data structure

Try it first...

## [NPV: Nested For Loop Example]{style="color:orange"} {.scrollable}

```{r npvfor, echo=TRUE, error=FALSE}

# write a function to compute npv
source("../R/compute_NPV.R")
compute_NPV
compute_NPV(20, discount=0.01, time=20)


# generate some input
damages = c(25,33,91,24)
damages
# sensitivity to discount rate
discount_rates = seq(from=0.01, to=0.04, by=0.005)
discount_rates
yr=10

# simple vectorization doesn't give all combinations -
# R recycles the shorter vector (gives 7 results with a warning, not 4x7=28)
compute_NPV(value=damages, discount=discount_rates, time=yr)

# compute npv's for all combinations of damages and discount rates
# first generate a dataframe to store results - rows = damages, columns = discount rates
npvs = data.frame(matrix(nrow=length(damages), ncol=length(discount_rates)))

# here we have a 2-dimensional array - rows and columns
head(npvs)

# now use a nested for loop to populate
for (i in 1:length(damages)) {
  for (j in 1:length(discount_rates)) {
    npvs[i,j] = compute_NPV(value=damages[i], discount=discount_rates[j], time=yr)
  }
}
npvs

# some data wrangling to make it pretty
colnames(npvs) = discount_rates
rownames(npvs) = damages
npvs

# how do I plot this with ggplot - add a column for original value and then rearrange
npvs$damage = damages
npvsg = npvs %>% pivot_longer(!damage, names_to="dis", values_to="npv")
head(npvsg)
ggplot(npvsg, aes(x=damage, y=npv, col=dis))+geom_point()+labs(col="Discount\nRate")

# how about summing all the damages
npv.total = npvsg %>% group_by(dis) %>% summarize(t=sum(npv))
ggplot(npv.total, aes(dis,t, fill=dis))+geom_col() + labs(x="Discount Rate", y="Total ($)")

```

## [An alternative method]{style="color:orange"} {.scrollable}

An alternative (R specific) is to use functions from the **purrr** package

Specifically *map* - which applies a function over a vector of inputs

```{r}
# use damages and discount rates from above
discount_rates
damages

# map applies compute_NPV for each discount rate,
# passing the full damages vector each time (compute_NPV is vectorized over value)
npvs = map(discount_rates, ~compute_NPV(value=damages, discount=.x, time=10))

# each element of npvs is a vector of NPVs (one per damage) at that discount rate
# convert to a data frame - each column is a discount rate
npvs = as.data.frame(npvs)
colnames(npvs) = discount_rates
npvs$damage = damages

# clean up and plot
npvsg = npvs %>% pivot_longer(!damage, names_to="dis", values_to="npv")
head(npvsg)
ggplot(npvsg, aes(x=damage, y=npv, col=dis))+geom_point()+labs(col="Discount\nRate")

# how about summing all the damages
npv.total = npvsg %>% group_by(dis) %>% summarize(t=sum(npv))
ggplot(npv.total, aes(dis,t, fill=dis))+geom_col() + labs(x="Discount Rate", y="Total ($)")
```
