---
title: 'Looping'
author: "Naomi Tague"
date: "Spring 2026"
format:
  revealjs:
    theme: solarized
    resources: ["images"]
editor: visual
execute:
  echo: true
  warning: true
  message: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE)

library(tidyverse)
library(purrr)
library(ggpubr)
```

Fixing Git when trying to pull from partner's repo

In terminal

*git config pull.rebase false*

Has to do with default settings

# <span style="color:orange">Looping (or repetition)

Repeating something is a fundamental concept in programming

We usually call this *looping*

All programming languages have ways to do this - sometimes hidden

Lets start with the basics of how you do repetition

# <span style="color:orange">For loops

**for** statement

-   defines a loop using a counter that is incremented each time you go through the loop

-   the counter - could be any variable (we often use *i*) but you could use what ever variable you want

-   repeats the commands between the *{* and *}* after the *for* keyword and loop definition

-   these commands are repeated until counter conditions is TRUE

```{r loop }

# simple example - use a loop to sum the numbers from 1 to 5
# a = 0+1+2+3+4+5
a=0
for (i in 1:5) {
 a = a+i
}
a

#if you want to keep track of a for each iteration
# start with a data structure to hold the results - there will be 5 iterations

a = rep(x=0, times=5)
for (i in 1:5) {
 a[i] = a[i]+i
}
a

# note that R has alternatives using function that does this looping for you
a = sum(seq(from=1, to=5))
a
```

# <span style="color:orange">Try a loop on your own

Make a loop that computes the mean/average of a set of numbers, use *runif* to generate a test/example set of 100 random numbers

Generating test data is a helpful way to check your code Take advantage of built in R functions to sample from distributions *normal*, *uniform*, ...

[Nice list of options](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Distributions.html)

```{r}
testnumbers = runif(min=0, max=20, n=100)
```

Check if your loop is correct using *mean*

# <span style="color:orange">My answer

```{r}
testnumbers = runif(min=0, max=20, n=100)

total = 0
nvalues = 100
for (i in 1:nvalues) {
  total = testnumbers[i]+total
}
total/nvalues

mean(testnumbers)
```

# <span style="color:orange">Another example

Calculating net income from revenue minus expenses

```{r}


# lets say we wanted to look income minus expenses for net income
# create some data for an example

income = runif(min=1000, max=5000, n=10)
income
expenses = rnorm(mean=1500, sd=500, n=10)
expenses

# it is possible that this approach could generate negative expenses
expenses = ifelse(expenses < 0, 0, expenses)


# normally we can just do
net = income-expenses
net


# lets make this a data frame for nice plotting
account = data.frame(income=income, expenses=expenses, net = net)
ggplot(account, aes(income, expenses))+geom_point()

# lets add a year
account$year = seq(from=2000, length.out=nrow(account))


# we could have computed net income with a for loop
# lets add a column for our net computed by a loop
# sometimes we use NA to show we haven't computed it yet, so "initialize" as NA

account$netloop=NA
account

# note I can use any variable I want to as a counter - 
# notice how I'm now using n to access each row of the data frame
for (n in 1:10) {
  account$netloop[n] = account$income[n]-account$expenses[n]
}

# as expected net and netloop are the same :)
head(account)
```

# <span style="color:orange">Why loop

While *sum* and *mean* are functions created by R community, what if you were doing something more complex to each row

What if we wanted to use data from multiple rows for each calculations?

For example, we want to carry forward net from the previous year (info from a different row)

So carry forward net income is current year's net income plus last year's net

```{r loop2 }

#View(account)

# note: we treat the first year differently because we don't have carry forward
# initialize our column to zero, start counter (n) at 2 instead of 1
# make the first value equal to the net of first year

account$carryforwardnet = 0
account$carryforwardnet[1]=account$net[1]
# now loop - looking backward one year to get previous years carry forward
for (n in 2:10) {
  account$carryforwardnet[n] = account$net[n]+account$carryforwardnet[n-1]
}

# plot - and use color to show current years contribution as positive or negative
ggplot(account, aes(year, carryforwardnet, fill=net, group=year))+geom_col()+scale_fill_gradient(low="red", high="black")


# another example - find the maximum speed from a set of speeds - lets make up an example
# by sampling from a random uniform distribution
speeds = runif(min=0, max=150, n=300)


maxspeed=0
for ( i  in 1:length(speeds)) {
  maxspeed = ifelse(speeds[i] > maxspeed, speeds[i], maxspeed)
}

maxspeed
max(speeds)



```

# <span style="color:orange">Add some complexity

Lets imagine that we have data on running speeds of a creature we are trying to develop a conservation plan for -

We think that when the creature is running at 35 km/hr and about they are hunting. We want to estimate how often they are hunting

Build a for loop to do this (there are other ways to get there with R but we want to practice looping)

As before, generate some test data using functions in R that allow you to generate data by sampling from distributions with known parameters

# My answer

```{r}

# make up sample data

speeds = runif(min=0, max=50, n=300)

# note how its useful to set a variable for our threshold hunting speed
# makes it easy to change and helps improve code readibility
hunting_speed = 35
num_hunting_speeds=0
for ( i  in 1:length(speeds)) {
  num_hunting_speeds = num_hunting_speeds + ifelse(speeds[i] >= hunting_speed, 1, 0)
}

num_hunting_speeds

# as a percent of recorded speeds
num_hunting_speeds/length(speeds)*100
```

# <span style="color:orange"> Steps for running function over multiple inputs

1.  design a data structure to store results: sometimes this is automatic but not always
2.  generate the data (or read it in)
3.  find ways to repeat things efficiently

-   *purr* family of function in R
-   more generally looping (*for*, *while*, ...others)

# <span style="color:orange"> A bit more on Looping

Loops can be "nested" with one loop inside the other

"Nesting" means loops inside loops

useful for "doing" something to each row and each column

**FOR** (i in 1:nrow) {

**FOR** (j in 1: ncol)

...

}}

In nested loops, the inner loop (last **FOR** in the sequences, gets done first)

So the loop above will all columns for the first row and then go to the next row

# <span style="color:orange"> Simple Example

Imagine we have a running speeds from our creature for different locations - stored in a array where rows are locations and columns are observations

Find number of times hunting occured across all locations

```{r}

# create some sample data
nobs = 100
nlocation = 15
# matrix function in R lets you initialize with data
creatureobs = matrix(runif(nobs*nlocation, min=0, max=100), nrow=nlocation)
# check to see that it a matrix with rows and columns
str(creatureobs)


# nested for loop example, use o as counter for observation; l as a counter for locatin
# outer loop for location
hunting_speed = 35
num_hunting_speeds=0
for (l in 1:nlocation) {
  # inner loop for observations 
  for (o in 1:nobs) {
     num_hunting_speeds = num_hunting_speeds + ifelse(creatureobs[l,o] >= hunting_speed, 1, 0)
  }
}

num_hunting_speeds
# as percent
num_hunting_speeds/(nobs*nlocation)*100
    
```

# <span style="color:orange"> More Complex Example

Example

Calculate NPV for

-   a range of different interest rates and
-   a range of damages
-   that may be incurred 10 years in the future

**Steps**

-   define inputs (interest rates, damages)
-   define output (NPV)
-   write the function
-   create a data structure to store results where we vary both interest rates and damages
-   use nested for loops to fill in the data structure

Try it first...

# <span style="color:orange">Rcode

```{r npvfor, echo=TRUE, error=FALSE}

# write a function to compute npv
source("../R/compute_NPV.R")
compute_NPV
compute_NPV(20, discount=0.01, time=20)


#generate some input
damages = c(25,33,91,24)
damages
# sensitivity to discount rate
discount_rates = seq(from=0.01, to=0.04, by=0.005)
discount_rates
yr=10

# note that the "simple" application doesn't work - why?
compute_NPV(20, discount=discount_rates, value=damages)

# compute some npv's for different discount rates for each damage
# first generate a dataframe to store results - we have both damages and discount rates that are changing - so we need rows and columns
npvs = data.frame(matrix(nrow=length(damages), ncol=length(discount_rates)))

# here we have a 2-dimensional array - rows and columns
head(npvs)

# now use a nested for loop to populate
 for (i in 1:length(damages)) {
         for (j in 1:length(discount_rates)) {
       npvs[i,j]= compute_NPV(value=damages[i],       discount=discount_rates[j],time=yr )

         }
 }
 npvs
 
 
 #some data wrangling to make it pretty
colnames(npvs)=discount_rates
rownames(npvs)=damages
 npvs
 
# how do I plot this with ggplot - add a column for original value and then rearrange
 npvs$damage = damages
 npvsg =npvs %>% pivot_longer(!damage,names_to="dis", values_to="npv")
 head(npvsg)
 ggplot(npvsg, aes(x=damage, y=npv, col=dis))+geom_point()+labs(col="Discount\nRate")
 
 # how about summing all the damages
 npv.total =npvsg %>% group_by(dis) %>% summarize(t=sum(npv))
 ggplot(npv.total, aes(dis,t, fill=dis))+geom_col() + labs(x="Discount Rate", y="Total ($)")

  
```

# An alternative method

An alternative (R specific) is to use functions from the **purr** package

Specifically *pmap* - which applies a function over multiple inputs

```{r}
 # we could also do the repetition with pmap from the purr family of functions

# use damages and discount rates from above

discount_rates
damages

# with purr family, pmap runs the function for each value and returns as a data frame 

npvs = pmap(list(discount_rates),compute_NPV, time=10, value=damages)
# turn into a data frame
npvs = as.data.frame(npvs, col.names=discount_rates)
colnames(npvs) = discount_rates
npvs$damage = damages
# clean up and plot
 npvsg =npvs %>% pivot_longer(!damage,names_to="dis", values_to="npv")
 head(npvsg)
 ggplot(npvsg, aes(x=damage, y=npv, col=dis))+geom_point()+labs(col="Discount\nRate")
 
 
 # how about summing all the damages
 npv.total =npvsg %>% group_by(dis) %>% summarize(t=sum(npv))
 ggplot(npv.total, aes(dis,t, fill=dis))+geom_col() + labs(x="Discount Rate", y="Total ($)")

 
 
```

# A bit more on nesting loops

we could even have multiple nesting (think about doings something to a 3 dimensional matrix time and space (x,y)

useful for "doings" something to each row and each column

```         
**FOR** (x in 1:nrow) {

    **FOR** (y in 1: ncol)  

      **FOR**  (t in 1: ntime)


            ...

    }
  }
}
```

To illustrate

```{r}

# lets ocean be a 3 dimensional array - space (x,y) and time
# lets say is a 2 by 3 spatial maps (2 units across, 3 units down), and 2 time steps
ocean = array(dim=c(20,30,10))

ocean

# lets make up data for our ocean

ocean[1,1,1] = runif(n=1, min=0, max=1)

# second time step
ocean[1,1,2] = runif(n=1, min=0, max=1)

# next column, first time step
ocean[2,1,1] = runif(n=1, min=0, max=1)

# for loops let us do this quickly
for (x in 1:nrow(ocean)) {
  for (y in 1:ncol(ocean)) {
    for (t in 1:dim(ocean)[3]) {
      ocean[x,y, t] = rnorm(mean=16+t, sd=2, n=1)
      # imagine running a model to estimate ocean temperature
      # then feeding it into a model that uses that temperature to impact corals
    }}}


```

# To Review for next class

*Looping.Rmd*

You can either clone the following repository or just go to the repo on github and download the file

[Example/Review Repo](https://github.com/naomitague/ESM_262_Examples.git)

"git clone https://github.com/naomitague/ESM_262_Examples.git"

**IF Your are Cloning**

-   make sure this is in a completely R project in a separate directory from your assignments directory (git doesn't like two different repos in the same place!)

-   find and review *Rmarkdown/starting_functions.Rmd* BEFORE next class - play with it; come ready to ask questions in class

If you just want the files

```{r error=TRUE}
library(devtools)
 install_github("naomitague/ESM_262_Examples")
```
