---
title: 'Looping'
author: "Naomi Tague"
date: last-modified
format:
  revealjs:
    theme: solarized
    resources: ["images"]
editor: visual
execute:
  echo: true
  warning: true
  message: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE)

library(tidyverse)
library(purrr)
library(ggpubr)
```

## [Add some complexity]{style="color:orange"}

Lets imagine that we have data on running speeds of a creature we are trying to develop a conservation plan for

We think that \* when the creature is running at 35 km/hr and above they are hunting. \* when they are moving below 5 km/hr they are socializing \* when they are moving as speeds between this they are traveling

We want to estimate how often they are in each of these categories based

Build a for loop to do this (there are other ways to get there with R but we want to practice looping)

As before, generate some test data using functions in R that allow you to generate data by sampling from distributions with known parameters

## [My answer]{style="color:orange"}

```{r}

# make up sample data
speeds = runif(min=0, max=50, n=300)

# setting a named variable for the threshold makes the code easier to read and change
hunting_speed = 35
social_speed = 5
num_hunting_speeds=0
num_social_speeds = 0
for ( i  in 1:length(speeds)) {
  num_hunting_speeds = num_hunting_speeds + ifelse(speeds[i] >= hunting_speed, 1, 0)
    num_social_speeds = num_social_speeds + ifelse(speeds[i] <  social_speed, 1, 0)
}

num_hunting_speeds
num_social_speeds

# other speeds are traveling, is remainder
num_traveling_speeds = length(speeds) - num_hunting_speeds - num_social_speeds

# as a percent of recorded speeds
num_hunting_speeds/length(speeds)*100
num_social_speeds/length(speeds)*100
num_traveling_speeds/length(speeds)*100

```

## [Steps for running a function over multiple inputs]{style="color:orange"}

1.  design a data structure to store results: sometimes this is automatic but not always
2.  generate the data (or read it in)
3.  find ways to repeat things efficiently

-   *purrr* family of functions in R
-   more generally looping (*for*, *while*, ...others)

## [A bit more on Looping]{style="color:orange"}

Loops can be "nested" with one loop inside the other

"Nesting" means loops inside loops

Useful for doing something to each row AND each column

```         
FOR (i in 1:nrow) {

  FOR (j in 1:ncol) {

    ...

  }
}
```

For each single step of the outer loop, the inner loop runs through **all** its iterations

So the loop above processes all columns for the first row, then moves to the next row

## [Simple Example]{style="color:orange"} {.scrollable}

Imagine we have running speeds from our creature for different locations - stored in a matrix where rows are locations and columns are observations

Find number of times hunting occurred across all locations

```{r}

# create some sample data
nobs = 100
nlocation = 15
# matrix function in R lets you initialize with data
creatureobs = matrix(runif(nobs*nlocation, min=0, max=100), nrow=nlocation)
# check to see that it is a matrix with rows and columns
str(creatureobs)


# nested for loop: use o as counter for observation; l as counter for location
# outer loop for location
hunting_speed = 35
num_hunting_speeds=0
for (l in 1:nlocation) {
  # inner loop for observations
  for (o in 1:nobs) {
     num_hunting_speeds = num_hunting_speeds + ifelse(creatureobs[l,o] >= hunting_speed, 1, 0)
  }
}

num_hunting_speeds
# as percent
num_hunting_speeds/(nobs*nlocation)*100

```

## [More Complex Example]{style="color:orange"}

Calculate NPV for

-   a range of different interest rates and
-   a range of damages
-   that may be incurred 10 years in the future

**Steps**

-   define inputs (interest rates, damages)
-   define output (NPV)
-   write the function
-   create a data structure to store results where we vary both interest rates and damages
-   use nested for loops to fill in the data structure

Try it first...

## [NPV: Nested For Loop Example]{style="color:orange"} {.scrollable}

```{r npvfor, echo=TRUE, error=FALSE}

# write a function to compute npv
source("../R/compute_NPV.R")
compute_NPV
compute_NPV(20, discount=0.01, time=20)


# generate some input
damages = c(25,33,91,24)
damages
# sensitivity to discount rate
discount_rates = seq(from=0.01, to=0.04, by=0.005)
discount_rates
yr=10

# simple vectorization doesn't give all combinations -
# R recycles the shorter vector (gives 7 results with a warning, not 4x7=28)
compute_NPV(value=damages, discount=discount_rates, time=yr)

# compute npv's for all combinations of damages and discount rates
# first generate a dataframe to store results - rows = damages, columns = discount rates
npvs = data.frame(matrix(nrow=length(damages), ncol=length(discount_rates)))

# here we have a 2-dimensional array - rows and columns
head(npvs)

# now use a nested for loop to populate
for (i in 1:length(damages)) {
  for (j in 1:length(discount_rates)) {
    npvs[i,j] = compute_NPV(value=damages[i], discount=discount_rates[j], time=yr)
  }
}
npvs

# some data wrangling to make it pretty
colnames(npvs) = discount_rates
rownames(npvs) = damages
npvs

# how do I plot this with ggplot - add a column for original value and then rearrange
npvs$damage = damages
npvsg = npvs %>% pivot_longer(!damage, names_to="dis", values_to="npv")
head(npvsg)
ggplot(npvsg, aes(x=damage, y=npv, col=dis))+geom_point()+labs(col="Discount\nRate")

# how about summing all the damages
npv.total = npvsg %>% group_by(dis) %>% summarize(t=sum(npv))
ggplot(npv.total, aes(dis,t, fill=dis))+geom_col() + labs(x="Discount Rate", y="Total ($)")

```

## [An alternative method]{style="color:orange"} {.scrollable}

An alternative (R specific) is to use functions from the **purrr** package

Specifically *map* - which applies a function over a vector of inputs

```{r}
# use damages and discount rates from above
discount_rates
damages

# map applies compute_NPV for each discount rate,
# passing the full damages vector each time (compute_NPV is vectorized over value)
npvs = map(discount_rates, ~compute_NPV(value=damages, discount=.x, time=10))

# each element of npvs is a vector of NPVs (one per damage) at that discount rate
# convert to a data frame - each column is a discount rate
npvs = as.data.frame(npvs)
colnames(npvs) = discount_rates
npvs$damage = damages

# clean up and plot
npvsg = npvs %>% pivot_longer(!damage, names_to="dis", values_to="npv")
head(npvsg)
ggplot(npvsg, aes(x=damage, y=npv, col=dis))+geom_point()+labs(col="Discount\nRate")

# how about summing all the damages
npv.total = npvsg %>% group_by(dis) %>% summarize(t=sum(npv))
ggplot(npv.total, aes(dis,t, fill=dis))+geom_col() + labs(x="Discount Rate", y="Total ($)")
```

## [A bit more on nesting loops]{style="color:orange"} {.scrollable}

We could even have multiple levels of nesting — think about doing something to a 3-dimensional matrix of time and space (x, y)

```         
FOR (x in 1:nrow) {

  FOR (y in 1:ncol) {

    FOR (t in 1:ntime) {

      ...

    }
  }
}
```

To illustrate

```{r}

# a 3-dimensional array: space (x=20, y=30) and time (10 steps)
ocean = array(dim=c(20,30,10))
str(ocean)

# fill with simulated temperature data using nested loops
for (x in 1:nrow(ocean)) {
  for (y in 1:ncol(ocean)) {
    for (t in 1:dim(ocean)[3]) {
      ocean[x,y,t] = rnorm(mean=16+t, sd=2, n=1)
      # imagine running a model to estimate ocean temperature
      # then feeding it into a model that uses that temperature to impact corals
    }
  }
}

```

## [To Review for next class]{style="color:orange"}

*Looping.Rmd*

You can either clone the following repository or just go to the repo on GitHub and download the file

[Example/Review Repo](https://github.com/naomitague/ESM_262_Examples.git)

"git clone https://github.com/naomitague/ESM_262_Examples.git"

**If you are Cloning**

-   make sure this is in a completely separate R project directory from your assignments directory (git doesn't like two different repos in the same place!)

-   find and review *Rmarkdown/starting_functions.Rmd* BEFORE next class - play with it; come ready to ask questions in class

If you just want the files

```{r error=TRUE}
library(devtools)
 install_github("naomitague/ESM_262_Examples")
```
