---
title: 'Testing'
format:
  revealjs:
    theme: solarized
    resources: ["images"]
editor: visual
execute:
  echo: true
  warning: true
  message: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE)
knitr::opts_chunk$set(error=TRUE)
#knitr::opts_knit$set(root.dir = '/Users/christina/Desktop/Courses.local/ESM262n/docs/packages')
library(tidyverse)
library(devtools)
library(roxygen2)
library(testthat)
library(here)


```

## [Testing]{style="color:orange"} {.scrollable}

Top 12 Reasons to Write Unit Tests - Burke and Coyner (Java programmers)

- Tests reduce bugs in new features
- Tests reduce bugs in existing features
- Tests defend against other programmers
- Testing forces you to slow down and think
- Testing makes development faster
- Tests reduce fear

Model developers now often use software to help them automate the testing process — reusing tests makes it efficient to repeat many checks as you develop and modify the code.

Particularly helpful when you have multiple modules (as in our mangrove example).

This type of software is available for R, Python, C, etc.

## [Insightful Effective Testing]{style="color:orange"} {.scrollable}

Good tests require you to *think* through what you want your code to do

- What should the output be for a given input?
- What should the output be for a range of inputs?
- What should the output be for edge cases (e.g. zero, negative, etc.)?

Requires you to come up with situations where you *know* what the answer should be — these make good tests.

- Doesn't have to be a specific number
  - Could be a range (e.g. NPV should be greater than zero)
  - Relationship between variables (e.g. max should be greater than min)
  - Relationship between outputs across inputs (e.g. if air temperature is always higher, snow should be less)

## [Insightful Effective Testing - More]{style="color:orange"} {.scrollable}

Good tests also require you to imagine the different ways your code might be used (and misused) and to test for those cases:

- Wrong input type, range, missing data, etc.
- Especially important when a function is part of a larger program

Good tests also require creativity — think of things that will break your code and test for those:

- What happens with outlier data (e.g. negative rainfall, or 1000mm of rain in a day)?

## [Testing in the AI Era]{style="color:orange"} {.scrollable}

Good testing becomes all the more important when AI is assisting you in writing code.

AI can be a great help in writing code, but it can also introduce errors if you are not careful.

AI can generate code that looks correct but may have hidden bugs or may not do exactly what you intended.

## [Formal Tests]{style="color:orange"} {.scrollable}

Most programming languages have ways for you to write tests and then run them automatically as part of a workflow:

- Makes testing much more efficient
- Allows you to run many tests at once
- Makes testing consistent across multiple developers

## [Testing in R]{style="color:orange"} {.scrollable}

In R, use the *testthat* library — install it and add it to your library calls.

If you are in the working directory where your functions are stored in a specific subdirectory (e.g. **R**), you can use *testthat* to automate testing:

- Need *devtools* and *testthat* libraries
- *test_dir("name")* — runs all tests in the "name" subdirectory (all files beginning with the word "test")
- *test_file("name")* — runs all the tests in a file called "name"

Testing in R uses **expectations**.

## [Expectations]{style="color:orange"} {.scrollable}

Expectations are functions you use to verify that your code is working as intended:

- `expect_true(condition)`
- `expect_equal(value, value)`
- `expect_match(string1, string2)`
- `expect_null(value)`
- `expect_length(object, n)`

```{r testingsimple, eval=TRUE, error=TRUE}

library(tidyverse)
library(devtools)
library(testthat)
# runs a test on a statement and checks to see if it is correct
# this one should work
expect_equal(2+2, 4)
# this one should 'throw an error'
expect_equal(2+5, 4)

# we can also use expect_true
# works with variables too
a = 200
expect_true(a > 2 )
a = 0
expect_true(a > 2 )

# matches
fish1 = "salmon"
fish2 = "trout"
expect_match(fish1, fish2)

# length
a = seq(from=1, to=10)
expect_length(a, 10)
expect_length(a, 11)
```

## [Testing Functions]{style="color:orange"} {.scrollable}

We can use expectations with functions to make sure they are working as expected.

Let's try with `compute_NPV`.

What do we know:

- If profit is zero, NPV should be zero
- If profit is positive, NPV should be positive


```{r, testfunctions}
library(here)

source(here("course-materials/R/compute_NPV.R"))
futurevalue = 0
expect_equal(compute_NPV(value=futurevalue, time=10, discount=0.01),0)
futurevalue = 100
expect_true(compute_NPV(value=futurevalue, time=10, discount=0.01) > 0)
futurevalue = -100
expect_true(compute_NPV(value=futurevalue, time=10, discount=0.01) < 0)

# lets introduce a bug in the function to see how tests catch it
compute_NPV = function(value, time, discount) {
  # use value to compute NPV
  result = value - (1 + discount)**time
  return(result)
}
expect_equal(compute_NPV(value=0, time=10, discount=0.01),0)

# lets reload the correct function
ex
```

## [Testing Functions - Syntax]{style="color:orange"} {.scrollable}

What if a function returns a list?

- You just access list items in your test
- You can also test for the names of the list items
- You can also test for relationships between list items (e.g. max should be greater than min)

Example using the *summary* function in R — imagine we are testing whether it works correctly.

Also explore generating test data for testing :)

```{r}

# generate some test data
testdata = rnorm(mean=25, sd=3, n=100)
summary(testdata)
# it is actually returning a named vector — what are the names?
names(summary(testdata))
# now we can access specific elements
summary(testdata)["Max."]
# and use in a test to make sure this function works
# the max should be greater than the min, right?

expect_true((summary(testdata)["Max."] - summary(testdata)["Min."]) > 0)
```

## Quick Check {.scrollable}

Write a test for the *mean* function in R:

- Generate test data
- Write a test using one of the `expect_*` functions
- Copy your answer to *Canvas Quiz*

## [Testing Functions - More Complex Example]{style="color:orange"} {.scrollable}

Let's create a function.

**Effective rainfall** — estimates rainfall that actually reaches the soil after accounting for interception by vegetation.

- Inputs: daily rainfall (`rain`, mm), interception capacity (`interceptC`, mm)
- Outputs: `totalrain` (mm), `totalraineff` (mm)

How it works: $raineff = \sum(rainfall - interceptC)$

## [Effective Rainfall — Defining and Testing]{style="color:orange"} {.scrollable}

```{r, eval=FALSE}


effective_rain=function(rain, interceptC=1) {
  
  # get total rain (in case user provides multiple days)
  totalrain = sum(rain)
  
  # interception happens every day so we need a new vector
  raineff = rain-interceptC
  totalraineff = sum(raineff)
  
  return(list(totalrain=totalrain, totalraineff=totalraineff))}


# simple tests
# if it never rains for 730 days, both totalrain and totalraineff should be zero
raindata = rep(0, times=730)
expect_that(effective_rain(raindata)$totalrain, equals(0))
# so far so good, now for effective rainfall

raindata = rep(0, times=730)
expect_that(effective_rain(raindata)$totalraineff, equals(0))

# what went wrong???

effective_rain=function(rain, interceptC=1) {
    
  # easy part - get total rain
  totalrain = sum(rain)
  
  # interception happens every day so we need to apply it per day
  # pmax ensures effective rain can't go below zero
  raineff = rain-interceptC
  raineff = pmax(raineff,0)
  totalraineff = sum(raineff)
  
  return(list(totalrain=totalrain, totalraineff=totalraineff))}

raindata = rep(0, times=730)
expect_that(effective_rain(raindata)$totalraineff, equals(0))

```

## [The Art of Testing]{style="color:orange"} {.scrollable}

It is good practice to create tests for *all* functions in a project — especially when there are multiple functions and/or multiple developers.

Recall when we designed function contracts at the beginning of class — testing is a way to make sure the program is honoring the contract.

You need to be creative:

- Test for things you know will work (quick way to find bugs)
- Usually we want to run multiple tests together

## [Testing in R — test_that]{style="color:orange"} {.scrollable}

A **test** runs one or more **expectation** functions and reports a message.

Format: `test_that(description, { series of expectations })`

```{r testing2, eval=FALSE}


source(here("course-materials/R/compute_NPV.R"))
# if profit is 0, NPV should be 0
# if time is 0, NPV should equal the original value
test_that("compute_NPV_works", {

expect_equal(compute_NPV(value=0,time=100,discount=0.1), 0)
expect_equal(compute_NPV(value=100, time=0, discount=0.1), 100)
})



```

## [Using Tests]{style="color:orange"} {.scrollable}

What we really want is to run tests *every time* we change something.

Store multiple tests in a *name.R* file and keep those files in the *tests/testthat* directory.

Two options:

- To run all tests in a file: `test_file("name")`
- To run all tests in a directory: `test_dir("name")` (also available from the Build menu)

**Goal:** Write tests for each function in your package/project.

Organization options:

- All tests for a given function in one file; all test files in one directory *(most common)*
- All test types in one file *(less common)*

## [Examples]{style="color:orange"} {.scrollable}

- `test_compute_NPV.R` — tests for `compute_NPV` function
- `test_spring_summary.R` — tests for `spring_summary` function

```{r testing, eval=FALSE}

# run tests in a file
test_file(here("course-materials/tests/test_compute_NPV.R"))

# run all tests in a directory 
test_dir(here("course-materials/tests"))
```

## Try On Your Own {.scrollable}

For one of the functions you have written:

- Write a test
- Run the test and make sure it works
- Add a *tests* directory to your project (Assignment 5)
- Add the test file there
- Run the test using *test_file* or *test_dir*

*Remember to use git add and commit to save your work.*

More expectation functions and testing approaches:

[testthat reference](https://testthat.r-lib.org/reference/index.html)
