
---
title: 'Functions'
format:
  revealjs:
    theme: solarized
    resources: ["images"]
editor: visual
execute:
  echo: true
  warning: true
  message: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE)
knitr::opts_chunk$set(error=TRUE)
#knitr::opts_knit$set(root.dir = '/Users/christina/Desktop/Courses.local/ESM262n/docs/packages')
library(tidyverse)
library(devtools)
library(roxygen2)
library(testthat)


```



# <span style="color:orange">Testing</span>

Top 12 Reasons to Write Unit Tests - Burke and Coyner (Java programmers)



* Tests reduce bugs in new features
* Tests reduce bugs in existing features
* Tests defend agains other programmers
* Testing forces you to slow down and think
* Testing makes development faster
* Tests reduce fear

Model developers now often use software to help them automate the testing process
Re-uses tests - makes it efficient to repeat many tests as you develop and modify the code 

Particularly helpful when you have multiple modules (as in our mangrove example)

This type of software is available for R, Python, C etc.

# <span style="color:orange">Testing in R</span>

In R, *testthat* library 
install and add this library

If you are in the working directory where you’ve stored the files in a specific directory (e.g **R**) you can use *testthat* to automate testing of your functions

* need devtools and “testthat” libraries
* test_dir(“name”) :runs all tests in the “name” subdirectory  (all files beginning with the word “test”
* test_file(“name”): runs all  the tests in a file called “name”

Testing in R uses expectation

# <span style="color:orange">Expectation</span>


tests you can use to make sure your code is working the way you think it should be working

* expect_true(condition)
* expect_equal(value, value)
* expect_match(string1, string2)
* expect_null(value)
* expect_length(object, n)



```{r testingsimple, eval=TRUE, error=TRUE}


# runs a test on a statement and checks to see if it is correct
# this one should work
expect_equal(2+2, 4)
# this one should 'throw and error'
expect_equal(2+5, 4)

# we can also use expect_true
# works with variables two
a = 200
expect_true(a > 2 )
a = 0
expect_true(a > 2 )

# matches
fish1 = "salmon"
fish2 = "trout"
expect_match(fish1, fish2)

# length
a = seq(from=1, to=10)
expect_length(a, 10)
expect_length(a, 11)


# We can use expectation with functions 
# and test just to make sure they are working as expected
source("R/compute_NPV.R")
futurevalue = 100
expect_true(compute_NPV(value=futurevalue, time=10, discount=0.01) > 0)
futurevalue = -100
expect_true(compute_NPV(value=futurevalue, time=10, discount=0.01) > 0)

# what if function returns a list?
# we can use the buit in summary function to see how this works
testdata = rnorm(mean=25, sd=3, n=100)
summary(testdata)
# it is actually returning a list, but what are the names?
names(summary(testdata))
# now we can specify them
summary(testdata)["Max."]
# and use in a test to make sure this function works -...hmm
# well the max should be greater than the min right?

expect_true((summary(testdata)["Max."]-summary(testdata)["Min."]) > 0)

# lets create a function with an error 
# imagine total recharge of water to soil after rain (in mm) is daily rainfall minus interception (in mm with default of 1mm) lost on any given day
# lets define a function that returns total rainfall and total recharge given an input of
# rainfall and then a value for interception

effective_rain=function(rain, interception_loss=1) {
  
  # easy part - get total rain
  totalrain = sum(rain)
  
  # interception happens everyday so we need a new vector
  recharge = rain-interception_loss
  totalrecharge = sum(recharge)
  
  return(list(totalrain=totalrain, totalrecharge=totalrecharge))}


# lets do some simple test
# if it never rains for 730 days then both should be zero
raindata = rep(0, times=730)
expect_that(effective_rain(raindata)$totalrain, equals(0))
# so far so good, now for recharge

raindata = rep(0, times=730)
expect_that(effective_rain(raindata)$totalrecharge, equals(0))

# what went wrong???

effective_rain=function(rain, interception_loss=1) {
  
  # easy part - get total rain
  totalrain = sum(rain)
  
  # interception happens everyday so we need a new vector
  recharge = rain-interception_loss
  recharge = pmax(recharge,0)
  totalrecharge = sum(recharge)
  
  return(list(totalrain=totalrain, totalrecharge=totalrecharge))}
raindata = rep(0, times=730)
expect_that(effective_rain(raindata)$totalrecharge, equals(0))

```

# The art of testing

It is good practice to create tests for functions (especially
useful if you are going to have multiple functions and/or multiple developers)

It make sure that the function returns what you think

Recall when we designed contracts at the beginning of class, testing is a way to make sure the program is honoring the contract

You need to be creative here - test for things you know will work (quick way to find bugs)

Usually we want to run multiple tests..

# <span style="color:orange">Testing in R</span>


A **test** runs an **expectation** function and returns a message

Format

test_that(description, series of expectations)
```{r testing2, eval=FALSE}


source("R/compute_NPV.R")
# lets do a simple test if profit is 0 , NPV should be 0
# if time is 0, NPV should be the original value
test_that("compute_NPV_works", {

expect_equal(compute_NPV(value=0,time=100,discount=0.1), 0)
expect_equal(compute_NPV(value=100, time=0, discount=0.1), 100)
})



```
# <span style="color:orange">Using tests </span>

What we really want is to have tests that we **always** run, everytime we change something


Store multiple test in a *name.R* file  store these in these in the *tests/testthat* directory

Two options

To run all tests in a file, use *test_file*


To run all tests in a directory,use *test_dir*  (you can also do this from Build menu)


Goal: Make tests for each function in your package 
We will talk more about making packages next week

```{r testing, eval=FALSE}

# put your test in a file
test_file("tests/test_compute_NPV.R")

# consider another function
source("R/spring_summary.R")
spring_summary

# to test this function we need to create some data


clim_data = as.data.frame(cbind(month=c(1:4), day=rep(1, times=4), year=c(2000,2000,2001,2001), 
                                 precip=rep(0, times=4), tavg=c(2,2,2,4)))
spring_summary(clim_data)
spring_summary(clim_data, spring_months=c(1,4))

# we can add data to a test
#save this to a file called test_spring_summary
test_that("spring_summary_works" ,
{clim_data = as.data.frame(cbind(month=c(1:4), day=rep(1, times=4), year=c(2000,2000,2001,2001), 
                                 precip=rep(0, times=4), tavg=c(2,2,2,4)))

expect_equal(spring_summary(clim_data, spring_months=4)$Pavg_spring,0)
expect_equal(spring_summary(clim_data, spring_months=4)$Tavg_spring, 4)
expect_equal(spring_summary(clim_data, spring_months=2)$Tavg_spring, 2)
expect_equal(spring_summary(clim_data, spring_months=c(1:4))$warmest_spring, 2001)

})

# see run a bunch of tests at once
test_file("tests/test_spring_summary.R")

# run all tests
test_dir("tests")
```

A good source for more expectation code and ways to do testing

[testing](https://testthat.r-lib.org/reference/index.html)
