---
title: 'Control Structures and Error Checking'
format:
  revealjs:
    theme: solarized
    resources: ["images"]
editor: visual
execute:
  echo: true
  warning: true
  message: true
---

```{r setup, include=FALSE}

library(tidyverse)

```

## <span style="color:orange"> What we've learned

- how to write a function
- how to generate data
- running functions with data

## <span style="color:orange"> Flow Control

Another KEY concept is flow control

Allowing your function to "do" different things depending on a conditions

CLASSIC example is

*IF* then *ELSE*

## If Else in R

Two ways in R

- *ifelse* function
  - ifelse(condition, TRUE, FALSE)
  - best used for simple application
- if..else structure
  - more formal way of deciding on two paths

## Example

classifying...

```{r, ifelse}
a = 20
classr  = ifelse(a > 90, "high","low")
classr

# also works with multiple values
a = runif(min=1, max=100, n=5)
a
# ifelse can work with vectors
classr  = ifelse(a > 90, "high","low")
classr

```

## <span style="color:orange"> Flow control in functions

best to use if {} else {}

see examaple in *summarize_diversity2.R*

```{r, ifelsefunction}

library(here)
source(here("course-materials/R/summarize_diversity2.R"))

flowers = sample(c("rose","peony","daisy"), replace=T, prob = c(0.1, 0.8, 0.1),size=30)

summarize_diversity(flowers, index="B")$simpson
summarize_diversity(flowers, index="A")$simpson

```

## <span style="color:orange"> Multiple Conditions {.scrollable}

If you have multiple conditions we can use *case_when*, requires *tidyverse* library - *switch* is another option

## <span style="color:orange"> Fishing Example {.scrollable}

Here's a simple example of how it works

Imagine you are recording fish catches at different locations (we are going to simulate this by sampling )

Fish have grades...

grade A, B, C

- steelhead are A,
- tuna are B
- everything else is C

R and other languages have ways to do this multiple alternatives flow control in R an example is

*case_when*

case_when(

condition \~ response,

condition \~ response ... )

First a simple example where we only have one fist

```{r was}
library(tidyverse)
fish = "steelhead"
case_when((fish =="steelhead")  ~ "A", (fish =="tuna") ~ "B", !(fish == "tuna") & !(fish=="steelhead") ~ "C")
```

## Make an ocean {.scrollable}

- notice how we build a function to make this 'cleaner'

```{r, ocean}
# apply to all of our ocean
# start by making a function - this will be cleaner!
classify_fish = function(fish) {
 class=  case_when((fish =="steelhead")  ~ "A", (fish =="tuna") ~ "B", !(fish == "tuna") & !(fish=="steelhead") ~ "C")
 return(class) }

classify_fish("steelhead")
classify_fish("cod")

# apply to our ocean
possible.fish = c("salmon","steelhead","shark","tuna","cod")
catch2 = base::sample(possible.fish, size=20, replace=T)
catch2

classify_fish(catch2)

```

## A more interesting example {.scrollable}

Lets imagine that we are monitoring pollution in a lake, and we want to write a function that will let us know (flag) if risk associated with nutrient pollution are high, medium or low

From ecological studies, we know that

*High Risk Conditions*

- risk is high if water temperature is greater than a threshold for more than 5 days, and
- mean nutrient concentration is greater than a high threshold

*Medium Risk Conditions*

- Risk is medium if water temperature is greater than a threshold for more than 5 days and
- mean nutrient concentration is greater than a medium threshold

*Everything else is low risk*

## Design the function {.scrollable}

**Inputs**:

- nutrient concentration for at least 5 days
- temperature for at least 5 days
- thresholds for temperature and nutrient (with default values)

**Output**:

- Mean Nutrient Concentration
- Pollution Risk as "low", "med" or "high"

Take a look at classify_lake

- example of while
- example of flow control with if..else and case_when

```{r, classifylake}

source(here("course-materials/R/classify_lake.R"))

# generate some data to try the function
temperature = runif(min=4, max=35, n=30)
nutrient = runif(min=5, max=40, n=30)

# try it
res=classify_lake(temperature=temperature, nutrient=nutrient)

# lets pick some numbers we know should give us a low value
nutrient = runif(min=0, max=5, n=30)
classify_lake(temperature=temperature, nutrient=nutrient)

# now try high
# use repeat to generate the same value multiple times
nutrient = rep(50, times=30)
temperature = rep(25, times=30)
classify_lake(temperature=temperature, nutrient=nutrient)
```

## <span style="color:orange"> Review. {.scrollable}

- We are using functions as a way to design modular programs (that might involve many functions that are linked together)

- We always start with clearly defininng our goal for the whole program and each piece (and diagramming this if needed for programs made of multiple functions)

- Always write your function in a text editor and then copy into R

- By convention we name files with functions in them by the name of the function.R e.g. **power_gen_orig.R**

- you can have R read a text file by source(“power_gen.R”) - make sure you are in the right working directory

- keep organized by keeping all functions in a subdirectory called *R*

- Eventually we will want our function to be part of a package (a library of many functions) - to create a package you must use this convention (name.R) place all function in a directory called “R”

## <span style="color:orange"> Error Checking {.scrollable}

What will your function do if user gives you garbage data, imagine that you have a function that computes microbial growth rates with temperature

- Two options

[1. error-checking (nice)]{style="color:skyblue"}

if temperature \< -100 or \> 100

- set results to NA
- or exit function with a warming

[2. assume user reads the contract :) return unrealistic values]{style="color:skyblue"}

if input -999.0, will still try to output growth rate

Error-checking is helpful if you are going to build a model made up of many functions- why?

## <span style="color:orange"> How to add error checking {.scrollable}

Consider this example (I'm going to put the function definition inline for now so you can easily see it - later we will save to our R subdirectory for reuse)

```{r eval=TRUE, echo=TRUE  }

power_gen = function(height, flow, rho=1000, g=9.8, Keff=0.8) {
  # make sure inputs are positive and if not set as NA so result will be NA
   if (height < 0) height=NA
  
 # an alterative is to exit the function prematurely 
  if (flow < 0) return("Flow cannot be less than 0")
  #    return("Height cannot be less than 0")
      
  # calculate power
    result = rho * height * flow * g * Keff
  return(result)
}

power_gen(2,3)
power_gen(-2,3)
power_gen(2,-3)
```
