---
title: 'Functions'
format:
  revealjs:
    theme: solarized
    resources: ["images"]
editor: visual
execute:
  echo: true
  warning: true
  message: true
---

```{r, setup, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
```

## Questions about functions so far

## Pair and Share


# Programming Always Starts with

[Conceptual Design]{style="color: cyan;"}

-   What is the goal
-   How can you break this goal into sub-task
-   For the overall project and each sub-task
    -   What are the inputs
    -   What are the outputs

# Modularity

-   Programs are made up of building blocks
-   A **main program** that controls the flow and **calls** functions/modules/building blocks


# <span style="color:blue"> Functions with factors

Lets generate a function that makes use of categorical data species diversity is a good example

"Simpson's Index (D) measures the probability that two individuals randomly selected from a sample will belong to the same species

Value is between 0 and 1, with lower values associated with *lower* diversity

See [Simpson Biodiversity](http://www.countrysideinfo.co.uk/simpsons.htm)

```{r diversity, echo=TRUE}

source(here("course-materials/R/compute_simpson_index.R"))
compute_simpson_index


possible.fish = as.factor(c("salmon","steelhead","shark","tuna","cod"))
# simulate a random recording of catch by fisherman


# note here is answer to above challenge
catch1 = sample(possible.fish, size=10, prob = c(0.2, 0.2, 0.1, 0.1, 0.4), replace=T)
# lets create a test case that should have low diversity, by repeating the same thing
catch2 = c(rep("salmon", times=10), rep("cod", times=10))

compute_simpson_index(catch1)
compute_simpson_index(catch2)

```


## Returning values from a function

return(value) - but what if you want to return multple items

Diversity and most frequent species?

In R - Lists

## List Data Type
## <span style="color:orange"> Lists \span {.scrollable}

-   Lists are the most “informal” data structures in R
-   List are really useful for keeping track of and organizing groups of things that are not all the same
-   A list could be a table where number of rows is different for each column
-   A list can have numeric, character, factors all mixed together
-   List are often used for returning more complex information from function (e.g. lm)

## Example of lists {.scrollable}

```{r introlist, echo=TRUE}

# make a list
sale = list(id=2, quality="high", contents=c("apple","cherry"), cost=c(4,5))
sale

#ways to access elements
sale$id
sale$what

# you can also access by list item number
# the [x] denotes the xth item in the list
sale[[3]]
sale[["contents"]]


# how do you get the second element in the vector that has the contents
# there are two ways


# add to a list
sale$location = "Farmers market"
sale
# or remove
sale$location = NULL
sale

# some tricky things
# correct accessing items in list
sale$cost
sale[[4]]

# works but
#sale[4]


sum(sale$cost)
sum(sale[[4]])

```


## For now - key parts of list

* Lists can contain different types of data (numeric, character, factors, vectors, data frames, other lists)

* In R to create a list
mylist = list(name="Bob", age=30, hobbies=c("fishing","hiking"), scores=data.frame(test1=c(90,80), test2=c(85,95)))

* to access something in a list - two options
mylist$name
mylist[["name"]]

There is more to say about lists but this is all we need for now

# Using lists to return multiple items from a function



We can use *lists* to return multiple,diverse pieces of information from our functions Lets start with diversity - many be want to know a bit more about the dataset

-   Simpson diversity
-   most frequent species
-   number of distinct species

Try it

```{r diversitylist, echo=TRUE}

# repeat with a list
source(here("course-materials/R/computediversity.R"))

computediversity

computediversity(catch1)
computediversity(catch2)
```

# Another example

We can also allow the user to make choices about what to return

```{r str, echo=TRUE}

source(here("course-materials/R/compute_season_meanflow.R"))
compute_season_flow

str = read.table(here("downloadables/Data", "str.txt"), header=T)

compute_season_flow(str)

compute_season_flow(str, kind="max")
```


## [Loops and Functions]{style="color:orange"}

Calculate NPV for

-   a range of different interest rates and
-   a range of damages
-   that may be incurred 10 years in the future

**Steps**

-   define inputs (interest rates, damages)
-   define output (NPV)
-   write the function
-   create a data structure to store results where we vary both interest rates and damages
-   use nested for loops to fill in the data structure

Try it first...

## [NPV: Nested For Loop Example]{style="color:orange"} {.scrollable}

```{r npvfor, echo=TRUE, error=FALSE}

# write a function to compute npv
source("../R/compute_NPV.R")
compute_NPV
compute_NPV(20, discount=0.01, time=20)


# generate some input
damages = c(25,33,91,24)
damages
# sensitivity to discount rate
discount_rates = seq(from=0.01, to=0.04, by=0.005)
discount_rates
yr=10

# simple vectorization doesn't give all combinations -
# R recycles the shorter vector (gives 7 results with a warning, not 4x7=28)
compute_NPV(value=damages, discount=discount_rates, time=yr)

# compute npv's for all combinations of damages and discount rates
# first generate a dataframe to store results - rows = damages, columns = discount rates
npvs = data.frame(matrix(nrow=length(damages), ncol=length(discount_rates)))

# here we have a 2-dimensional array - rows and columns
head(npvs)

# now use a nested for loop to populate
for (i in 1:length(damages)) {
  for (j in 1:length(discount_rates)) {
    npvs[i,j] = compute_NPV(value=damages[i], discount=discount_rates[j], time=yr)
  }
}
npvs

# some data wrangling to make it pretty
colnames(npvs) = discount_rates
rownames(npvs) = damages
npvs

# how do I plot this with ggplot - add a column for original value and then rearrange
npvs$damage = damages
npvsg = npvs %>% pivot_longer(!damage, names_to="dis", values_to="npv")
head(npvsg)
ggplot(npvsg, aes(x=damage, y=npv, col=dis))+geom_point()+labs(col="Discount\nRate")

# how about summing all the damages
npv.total = npvsg %>% group_by(dis) %>% summarize(t=sum(npv))
ggplot(npv.total, aes(dis,t, fill=dis))+geom_col() + labs(x="Discount Rate", y="Total ($)")

```

## [An alternative method]{style="color:orange"} {.scrollable}

An alternative (R specific) is to use functions from the **purrr** package

Specifically *map* - which applies a function over a vector of inputs

```{r}
# use damages and discount rates from above
discount_rates
damages

# map applies compute_NPV for each discount rate,
# passing the full damages vector each time (compute_NPV is vectorized over value)
npvs = map(discount_rates, ~compute_NPV(value=damages, discount=.x, time=10))

# each element of npvs is a vector of NPVs (one per damage) at that discount rate
# convert to a data frame - each column is a discount rate
npvs = as.data.frame(npvs)
colnames(npvs) = discount_rates
npvs$damage = damages

# clean up and plot
npvsg = npvs %>% pivot_longer(!damage, names_to="dis", values_to="npv")
head(npvsg)
ggplot(npvsg, aes(x=damage, y=npv, col=dis))+geom_point()+labs(col="Discount\nRate")

# how about summing all the damages
npv.total = npvsg %>% group_by(dis) %>% summarize(t=sum(npv))
ggplot(npv.total, aes(dis,t, fill=dis))+geom_col() + labs(x="Discount Rate", y="Total ($)")
```
