---
title: 'Functions'
format:
  revealjs:
    theme: solarized
    resources: ["images"]
editor: visual
execute:
  echo: true
  warning: true
  message: true
---

```{r, setup, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
```

## [Loops and Functions]{style="color:orange"}

Calculate NPV for

- a range of different interest rates and
- a range of damages
- that may be incurred 10 years in the future

**Steps**

- define inputs (interest rates, damages)
- define output (NPV)
- write the function
- create a data structure to store results where we vary both interest rates and damages
- use nested for loops to fill in the data structure

Try it first...

## [NPV: Nested For Loop Example]{style="color:orange"} {.scrollable}

```{r npvfor, echo=TRUE, error=FALSE}

# write a function to compute npv
source("../R/compute_NPV.R")
compute_NPV
compute_NPV(20, discount=0.01, time=20)


# generate some input
damages = c(25,33,91,24)
damages
# sensitivity to discount rate
discount_rates = seq(from=0.01, to=0.04, by=0.005)
discount_rates
yr=10

# simple vectorization doesn't give all combinations -
# R recycles the shorter vector (gives 7 results with a warning, not 4x7=28)
compute_NPV(value=damages, discount=discount_rates, time=yr)

# compute npv's for all combinations of damages and discount rates
# first generate a dataframe to store results - rows = damages, columns = discount rates
npvs = data.frame(matrix(nrow=length(damages), ncol=length(discount_rates)))

# here we have a 2-dimensional array - rows and columns
head(npvs)

# now use a nested for loop to populate
for (i in 1:length(damages)) {
  for (j in 1:length(discount_rates)) {
    npvs[i,j] = compute_NPV(value=damages[i], discount=discount_rates[j], time=yr)
  }
}
npvs

# some data wrangling to make it pretty
colnames(npvs) = discount_rates
rownames(npvs) = damages
npvs

# how do I plot this with ggplot - add a column for original value and then rearrange
npvs$damage = damages
npvsg = npvs %>% pivot_longer(!damage, names_to="dis", values_to="npv")
head(npvsg)
ggplot(npvsg, aes(x=damage, y=npv, col=dis))+geom_point()+labs(col="Discount\nRate")

# how about summing all the damages
npv.total = npvsg %>% group_by(dis) %>% summarize(t=sum(npv))
ggplot(npv.total, aes(dis,t, fill=dis))+geom_col() + labs(x="Discount Rate", y="Total ($)")

```

## [An alternative method]{style="color:orange"} {.scrollable}

An alternative (R specific) is to use functions from the **purrr** package

Specifically *map* - which applies a function over a vector of inputs

```{r}
# use damages and discount rates from above
discount_rates
damages

# map applies compute_NPV for each discount rate,
# passing the full damages vector each time (compute_NPV is vectorized over value)
npvs = map(discount_rates, ~compute_NPV(value=damages, discount=.x, time=10))

# each element of npvs is a vector of NPVs (one per damage) at that discount rate
# convert to a data frame - each column is a discount rate
npvs = as.data.frame(npvs)
colnames(npvs) = discount_rates
npvs$damage = damages

# clean up and plot
npvsg = npvs %>% pivot_longer(!damage, names_to="dis", values_to="npv")
head(npvsg)
ggplot(npvsg, aes(x=damage, y=npv, col=dis))+geom_point()+labs(col="Discount\nRate")

# how about summing all the damages
npv.total = npvsg %>% group_by(dis) %>% summarize(t=sum(npv))
ggplot(npv.total, aes(dis,t, fill=dis))+geom_col() + labs(x="Discount Rate", y="Total ($)")
```
## # <span style="color:orange"> Lets practice:: Automobile Power Generation </span>


One of the equations used to compute automobile fuel efficiency is given below. This is the power required to keep a car moving at a given speed, and is as follows:


$Pb = crolling * m *g *V + \frac{1}{2} * A*p_{air}*c_{drag}*V^3$


where crolling and cdrag are rolling and aerodynamic resistive coefficients, typical values are 0.015 and 0.3, respectively.

- V: is vehicle speed (assuming no headwind) in m/s (or mps) m: is vehicle mass in kg

- A is surface area of car $(m^2)$

- *g*: is acceleration due to gravity $(9.8 m/s^2 )$ 

- p_air = density of air $(1.2kg/m^3 )$

- Pb is power in Watts

What would be the function's contract - input/parameters? what will it do?

# <span style="color:orange"> Lets practice:: Automobile Power Generation </span>

1. Write a function to compute power, given a truck of m=31752 kg (parameters for a heavy truck) for a range of different highway speeds, the surface area of this truck  is 16 $m^2$

* 30 m/s
* 80 km/hr
* 120 km/hr
* 40 km/hr
* 5 km/hr


2. Plot power as a function of speed

3. Recompute for a lighter vehicle (15,000 kg) 
Add a second line to your graph for lighter vehicle

* Keep your function in its own file, store it in a subdirectory called *R* and then create a **separate** Rmarkdown (or Quarto) document to use the function and store the Rmarkdown in a different directory

# **Good Programming Practices**

Name variables such that they are easy to understand

Describe the contract at the top of the function (following the format for the *power_gen_orig.R* example)

Also include the comments in the Quarto 