---
title: 'Looping'
author: "Naomi Tague"
date: last-modified
format:
  revealjs:
    theme: solarized
    resources: ["images"]
editor: visual
execute:
  echo: true
  warning: true
  message: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE)

library(tidyverse)
library(purrr)
library(ggpubr)
```

## Quick Review: Creating strings with sprintf. {.scrollable}

**sprintf** used something called *format string syntax* to create formatted strings

*format string syntax* is used in many places -for searching for and creating strings - AND - its used by many programming languages

*sprintf* uses format 'code' to tell it how to create a string, based on the values you give it

Example

```{r}
sprintf("We found %d %s %3.1f miles from shore", 4,"Otters", 18.95632342)
```

-   **%s** replace with a string
-   **%d** replace with an integer value
-   **%f** replace with a real value
-   **%4.1f** replace with a real value with 4 digits, two after decimal
-   **\\n** add a line return

## Example {.scrollable}

make a string with **sprintf** and add to plot title

```{r, eval=TRUE, echo=TRUE}
# # create vector of possible flowers 
possible.flowers = c("tulip","rose","lily","iris","marigold","daisy")

# we can use sample to simulate a random recording of purchases at the market, lets create data that had 300 flower purchases

purchases = sample(possible.flowers, size=20, replace=T)
purchases= as.factor(purchases)

# make it a data frame for plotting later
# maybe we want to add more information about the purchases later - like price, time of day, etc. - so make column name flower
purchases  = data.frame(flower=purchases) 

# to quickly get information of flower purchases 
summary(purchases$flower)
# sprintf creates a string %s mean use what ever is after the , to fill in a string
plottitle=sprintf("The most frequently bought flower was %s\n There were %d bought", names(which.max(summary(purchases$flower))), max(summary(purchases$flower)))

# now plot
ggplot(purchases, aes(flower, fill=flower))+geom_histogram(stat="count")+labs(title=plottitle)+
  theme(legend.position="none")

```

## <span style="color:orange"> What we've explored \span {.scrollable}

-   common data types
-   ways to generate data
-   string format syntax

## [Looping (or repetition)]{style="color:orange"}

Repeating something is a fundamental concept in programming

We usually call this *looping*

All programming languages have ways to do this - sometimes hidden

Lets start with the basics of how you do repetition - \* the *for* loop - used in many languages

## [For loops]{style="color:orange"} {.scrollable}

**for** statement

-   defines a loop using a counter that is incremented each time you go through the loop

-   the counter - could be any variable (we often use *i*) but you could use whatever variable you want

-   repeats the commands between the *{* and *}* after the *for* keyword and loop definition

-   these commands are repeated until the counter condition is TRUE

```{r loop }

# simple example - use a loop to sum the numbers from 1 to 5
# a = 0+1+2+3+4+5
a=0
for (i in 1:5) {
 a = a+i
}
a
```

## Break it down

-   for (i in 1:5) - mean you start with i=1

-   the commands in the loop are executed - a = a+i

-   then you go back to the top of the loop and increment i to 2

-   i = 1

-   a = 0 + 1 = 1

-   i = 2

-   a = 1 + 2 = 3

-   i = 3

-   a = 3 + 3 = 6

-   i = 4

-   a = 6 + 4 = 10

-   i = 5

-   a = 10 + 5 = 15

after the loop a = 15

## Keeping track of what happens each step/interation

-   start with a data structure to hold the results -

-   initialize to 0

    -   *its good practice to initialize data before a loop*
    -   initialize means to create a data structure and set values, usually to zero but sometimes NA

```{r}

# if you want to keep track of a for each iteration
# start with a data structure to hold the results - 
# there will be 5 iterations, so a vector of length 5
# initialize to 0 

av = rep(x=0, times=5)

# now we can fill in the vector with the sum at each iteration
# we can use the counter (i) to access the correct element of the vector
# note that we have to treat the first iteration differently because we don't have a previous value to add to

for (i in 1:5) {
  if (i==1) {
    av[i] = 1
  } else {
  av[i] = av[i-1]+i
  }
}
av

# note that R has alternatives using functions that do this looping for you
a = sum(seq(from=1, to=5))
a
```

## [When to use loops]{style="color:orange"}

-   When you want to repeat something a specific number of times
-   R has many built in functions to do this
    -   BUT sometimes won't do exactly what you want
    -   other languages also use versions of the *for* loop

## [Try a loop on your own]{style="color:orange"} {.scrollable}

Make a loop that computes

-   the mean of a set of numbers,
-   use *runif* to generate a test/example set of 100 random numbers
-   test your loop by comparing to the built in *mean* function

Generating test data is a helpful way to check your code. Take advantage of built-in R functions to sample from distributions as we did last class

```{r}
testnumbers = runif(min=0, max=20, n=100)
```

Check if your loop is correct using *mean*

## [My answer]{style="color:orange"}

```{r}
testnumbers = runif(min=0, max=20, n=100)

total = 0
nvalues = 100
for (i in 1:nvalues) {
  total = testnumbers[i]+total
}
total/nvalues

mean(testnumbers)
```

## [Another example]{style="color:orange"} {.scrollable}

Lets look at some more slightly more complex examples

Calculating net income from revenue minus expenses

```{r}

# lets say we wanted to look at income minus expenses for net income
# create some data for an example

income = runif(min=1000, max=5000, n=10)
income
expenses = rnorm(mean=1500, sd=500, n=10)
expenses

# it is possible that this approach could generate negative expenses
expenses = ifelse(expenses < 0, 0, expenses)

# lets make this a data frame for organization
account = data.frame(income=income, expenses=expenses)
# lets add a year
account$year = seq(from=2000, length.out=nrow(account))

# now compute net income - we could do this with a loop but R is designed to do this without loops
# normally we can just do
account = account %>% mutate(net = income-expenses)
head(account)


# lets plot
ggplot(account, aes(income, expenses))+geom_point()



# we could have computed net income with a for loop
# lets add a column for our net computed by a loop
# sometimes we use NA to show we haven't computed it yet, so "initialize" as NA

account$netloop=NA
account

# note that I can use any variable I want as a counter -
# notice how I'm now using n to access each row of the data frame
for (n in 1:10) {
  account$netloop[n] = account$income[n]-account$expenses[n]
}

# as expected net and netloop are the same :)
head(account)
```

## [Why loop: cross-row dependencies]{style="color:orange"} {.scrollable}

While *sum* and *mean* are functions created by the R community, what if you were doing something more complex to each row?

What if we needed data from a **previous row** for each calculation?

Example: carry-forward net income = current year's net + last year's carry-forward

```{r loop2 }

# note: we treat the first year differently because we don't have carry forward
# initialize our column to zero, start counter (n) at 2 instead of 1
# make the first value equal to the net of first year

account$carryforwardnet = 0
account$carryforwardnet[1]=account$net[1]
# now loop - looking backward one year to get previous years carry forward
for (n in 2:10) {
  account$carryforwardnet[n] = account$net[n]+account$carryforwardnet[n-1]
}

# plot - use color to show current year's contribution as positive or negative
ggplot(account, aes(year, carryforwardnet, fill=net, group=year))+geom_col()+
  scale_fill_gradient(low="red", high="black")
```

## [Why loop: sequential comparisons]{style="color:orange"} {.scrollable}

Another case where loops are necessary: comparing across elements one at a time

Lets say we are measuring speeds and we want to determine the number of times speeds are above a threshold

There are many ways to do this - but *for* will always work

```{r}

# speed limit 
speedthresh = 120
# create some sample speeds 
speeds = runif(min=0, max=150, n=300)

nthresh = 0
for ( i  in 1:length(speeds)) {
   if(speeds[i] > speedthresh)
     nthresh = nthresh + 1
}

nthresh

```

## [Add some complexity]{style="color:orange"} {.scrollable}

Lets imagine that we have data on running speeds of a creature we are trying to develop a conservation plan for

We think that

-   when the creature is running at 35 km/hr and above they are hunting.
-   when they are moving below 5 km/hr they are socializing
-   when they are moving as speeds between this they are traveling

We want to estimate how often they are in each of these categories based

Build a for loop to do this (there are other ways to get there with R but we want to practice looping)

As before, generate some test data using functions in R that allow you to generate data by sampling from distributions with known parameters

## [My answer]{style="color:orange"}

```{r}

# make up sample data
speeds = runif(min=0, max=50, n=500)

# setting a named variable for the threshold makes the code easier to read and change
hunting_speed = 35
social_speed = 5
num_hunting_speeds=0
num_social_speeds = 0
for ( i  in 1:length(speeds)) {
  num_hunting_speeds = num_hunting_speeds + ifelse(speeds[i] >= hunting_speed, 1, 0)
    num_social_speeds = num_social_speeds + ifelse(speeds[i] <  social_speed, 1, 0)
}


for (i in 1:300)
num_hunting_speeds
num_social_speeds

# other speeds are traveling, is remainder
num_traveling_speeds = length(speeds) - num_hunting_speeds - num_social_speeds

# as a percent of recorded speeds
num_hunting_speeds/length(speeds)*100
num_social_speeds/length(speeds)*100
num_traveling_speeds/length(speeds)*100

```

# See Assignment 3
