---
title: 'Functions'
format:
  revealjs:
    theme: solarized
    resources: ["images"]
editor: visual
execute:
  echo: true
  warning: true
  message: true
---

```{r, setup, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
```

## Programming Always Starts {style="color:orange"} {.scrollable}

- What is the goal
- How can you break this goal into sub-task
- For the overall project and each sub-task
  - What are the inputs
  - What are the outputs

## Modularity {.scrollable}

Programs are made up of building blocks

When you start a programming/data science project - you should begin by:

- Breaking your goal down into individual pieces

- Identifying instructions that can be reused

- Identify tasks that might be useful to others

- Identify tasks that you might want to be able to substitute different methods for

## Questions about functions so far {.scrollable}

<span style="color:blue"> Homework

Lets generate a function that makes use of categorical data species diversity is a good example

Diversity (D) measures the probability that two individuals randomly selected from a sample will belong to the same species

To get value is between 0 and 1, with lower values associated with *lower* diversity 1-D from Assignment (Simpson Index)

## Pair and Share

## My version {.scrollable}

```{r diversity, echo=TRUE}

source(here("course-materials/R/compute_diversity.R"))
compute_diversity


possible.fish = as.factor(c("salmon","steelhead","shark","tuna","cod"))
# simulate a random recording of catch by fisherman

# we can control sample probablities as long as it sums to 1 - here we have a mix of fish but cod is most common

# to do this we assign probabities to each fish 
# high prob more likely to be sampled
# probabilities must sum to 1 
catch1 = sample(possible.fish, size=10, prob = c(0.2, 0.2, 0.1, 0.1, 0.4), replace=T)

# lets create a test case that should have low diversity, making most fish salmon
catch2 = sample(possible.fish, size=10, prob = c(0.8, 0.05, 0.05, 0.05, 0.05), replace=T)

# note that the diversity index works 'backwards' - higher values are less diverse 
compute_diversity(catch1)
compute_diversity(catch2)

```

## Returning values from a function {.scrollable}

return(value) - but what if you want to return multple items

Diversity and most frequent species?

In R -We use Lists

## <span style="color:orange"> Lists \span {.scrollable}

- Lists are the most “informal” data structures in R
- List are really useful for keeping track of and organizing groups of things that are not all the same
- A list could be a table where number of rows is different for each column
- A list can have numeric, character, factors all mixed together
- List are often used for returning more complex information from function (e.g. lm)

## Example of lists {.scrollable}

```{r introlist, echo=TRUE}

# make a list
sale = list(id=2, quality="high", contents=c("apple","cherry"), cost=c(4,5))
sale

#ways to access elements
sale$id
sale$what

# you can also access by list item number
# the [x] denotes the xth item in the list
sale[[3]]
sale[["contents"]]


# how do you get the second element in the vector that has the contents
# there are two ways


# add to a list
sale$location = "Farmers market"
sale
# or remove
sale$location = NULL
sale

# some tricky things
# correct accessing items in list
sale$cost
sale[[4]]

# works but
#sale[4]


sum(sale$cost)
sum(sale[[4]])

```

## For now - key parts of list {.scrollable}

- Lists can contain different types of data (numeric, character, factors, vectors, data frames, other lists)

- In R to create a list mylist = list(name="Bob", age=30, hobbies=c("fishing","hiking"), scores=data.frame(test1=c(90,80), test2=c(85,95)))

- to access something in a list - two options mylist\$name mylist\[\["name"\]\]

There is more to say about lists but this is all we need for now

## Using lists to return multiple items from a function {.scrollable}

We can return a list from a function to give the user multiple pieces of information

```{r, simple}

# a function that returns the max and min of a vector of numbers
minmax = function(x) {
  minx = min(x)
  maxx = max(x)
  return(list(min=minx, max=maxx))
}

minmax(c(3,5,2,8,1))


```

## More complex example {.scrollable}

Lets say we want to compute a water quality index that is based on multiple measurements (e.g. temperature, dissolved oxygen, turbidity)

- check if each measurement is above, within, or below a thresholdrange
- return rating for each measurement (e.g. good, moderate, poor if above, within, or below threshold)
- return an overall numeric rating
- return an overall index of water quality (e.g. good, moderate, poor)

Look through function - and ask you neighbor any questions

```{r, usefunction, echo=TRUE}

# Source the function
source(here("course-materials/R/analyze_water_quality.R"))

# --- Use defaults (most common case) ---
results <- analyze_water_quality(ph = 7.2, dissolved_oxygen = 8.5, turbidity = 2.1)
results

# --- Override one limit, e.g. stricter turbidity standard for drinking water ---
results_strict <- analyze_water_quality(
  ph = 7.2, dissolved_oxygen = 8.5, turbidity = 2.1,
  turbidity_max = 1.0
)
results_strict

```

## Practice

Modify your diversity function to return a list with

- Simpson diversity
- most frequent species
- number of distinct species

Try it

## My answer {.scrollable}

```{r diversitylist, echo=TRUE}

library(here)
library(tidyverse)
# repeat with a list
source(here("course-materials/R/summarize_diversity.R"))

# lets compare two catches - with different levels of diversity
possible.fish =c("salmon","steelhead","shark","tuna","cod")

# we can control sample probablities as long as it sums to 1 - here we have a mix of fish but cod is most common

# to do this we assign probabities to each fish 
# high prob more likely to be sampled
# probabilities must sum to 1 
catch1 = sample(possible.fish, size=10, prob = c(0.2, 0.2, 0.1, 0.1, 0.4), replace=T)

# lets create a test case that should have low diversity, making most fish steelhead
catch2 = sample(possible.fish, size=10, prob = c(0.1, 0.7, 0.1, 0.1, 0.1), replace=T)

summarize_diversity

# run function and save results to a variable
d1 = summarize_diversity(catch1)
d2 = summarize_diversity(catch2)
d1
d2

# just index
d1$simpson
d2$simpson
```

## Other ways to use functions you might not think of {.scrollable}

- Use parameters to allow the user to make choices about what to return

- Use parameters to allow the user to make choices about how to compute something

- Use parameters to allow the user to make choices about what data to use

- Use parameters to apply different weighting to an index calculation

## Example {.scrollable}

Lets let user choose between two different ways to compute the index

and also whether to show a plot of the data or not

```{r parametercontrol, echo=TRUE, eval=TRUE}

library(here)
library(tidyverse)
source(here("course-materials/R/summarize_diversity2.R"))

testfish = c("cod","trout","salmon")
# create a test catch that is dominated by cod, but has some trout and salmon
testcatch = sample(testfish, size=20, replace=T, prob=c(0.8,0.2,0.2))

# lets compare different ways to compute the index  
d1 = summarize_diversity(testcatch, index="A")
d2 = summarize_diversity(testcatch, index="B")

# just look at index
d1$simpson
d2$simpson

# run with instruction to plot
d3 = summarize_diversity(testcatch, index="B", showplot=TRUE)
d3$plt
```
