---
title: 'Functions'
format:
  revealjs:
    theme: solarized
    resources: ["images"]
editor: visual
execute:
  echo: true
  warning: true
  message: true
---

```{r, setup, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
```

# Programming Always Starts with

[ Conceptual Design ]{style="color: cyan;"}

-   What is the goal
-   How can you break this goal into sub-task
-   For the overall project and each sub-task
    -   What are the inputs
    -   What are the outputs

# Modularity

-   Programs are made up of building blocks
-   A **main program** that controls the flow and **calls** functions/modules/building blocks

[Modular Housing](https://images.app.goo.gl/moGdSXVHWiRvHm3X8/)

# Modules take you from inputs to outputs in pieces

-   helpful to start a software engineering project by **diagramming** or **flow charting** your program

-   what is the goal

-   what are key building blocks - modules or functions

-   inputs and output

-   how do you put the blocks together to go from input to output - main program

# <span style="color:blue"> Functions with factors 

Lets generate a function that makes use of categorical data
species diversity is a good example

"Simpson's Index (D) measures the probability that two individuals randomly selected from a sample will belong to the same species 

Value is between 0 and 1, with lower values associated with *lower* diversity

See 
[Simpson Biodiversity](http://www.countrysideinfo.co.uk/simpsons.htm)


```{r diversity, echo=TRUE}

source("R/compute_simpson_index.R")
compute_simpson_index


possible.fish = as.factor(c("salmon","steelhead","shark","tuna","cod"))
# simulate a random recording of catch by fisherman


# note here is answer to above challenge
catch1 = sample(possible.fish, size=10, prob = c(0.2, 0.2, 0.1, 0.1, 0.4), replace=T)
# lets create a test case that should have low diversity, by repeating the same thing
catch2 = c(rep("salmon", times=10), rep("cod", times=10))

compute_simpson_index(catch1)
compute_simpson_index(catch2)

```


# Using lists to return multiple items from a function

We can use *lists* to return multiple,diverse pieces of information from our functions
Lets start with diversity - many be want to know a bit more about the dataset

* Simpson diversity
* most frequent species
* number of distinct species



```{r diversitylist, echo=TRUE}

# repeat with a list
source("R/computediversity.R")

computediversity

computediversity(catch1)
computediversity(catch2)
```


# Another example

We can also allow the user to make choices about what to return

```{r str, echo=TRUE}

source("R/compute_season_meanflow.R")
compute_season_flow

str = read.table("data/str.txt", header=T)
compute_season_flow(str)

compute_season_flow(str, kind="max")
```

