---
title: 'Git Part 3 '
format:
  revealjs:
    theme: solarized
    resources: ["images"]
editor: visual
execute:
  echo: true
  warning: true
  message: true
---

```{r, setup, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
library(here)
library(tidyverse)
```

## Any Questions?

## Git

![git conflicts](images/week2/git_workflow_diagram.svg)

## Conflicts Review {.scrollable}

* note that conflicts must be resolved before continuing

* remember if you really mess up you can *revert* to earlier version

* frequent *push* and *pull* can help avoid conflicts

## Branching 

Sometimes you want to experiment with a new feature or make a change that might break things, could get messy to do this with *commit* and then *revert* if it doesn't work

*Git* thought of this and created *branching* 

*Branching* allows you to create a separate version of your code where you can make changes without affecting the main version (called *main*  branch)

Later once branch is working you can *merge* into *main* branch

## Branching

![git branchings](images/week3/git_branching_animals.svg)

## Lets try

* working in your class repo

* create a new branch called *experiment* 


* add a new quarto document, and in it describe some data analysis that you'd like to experiment with


## Merging

* in a collaborative project, or if you want to maintain a good copy of your code, you might want to merge branch back into main

* go back to *main* branch on git
* *pull* to get any recent changes
* click *merge* and select the branch you want to merge in

or in terminal 

*git merge experiment*

Try it

## Forking

Forking is more removed than branching - it creates a copy of the entire repository under your GitHub account

Forks are their own *repo*

How do forks interact with original
  - fork can pull from original repo
  - add a *pull request* to add changes from *fork* to orginal
  
Forks are good when you are working on a repo that you are not closely collaborating with others  
(and the owner of the repo maintains strong controll over it)


We won't practice them here but good to know


## Recall useful Git commands in Terminal {.scrollable}

-   `git status` - shows you the status of your files (which ones are changed, which ones are staged for commit, etc.)

-   `git log` - shows you the commit history with commit messages and IDs

-   `git diff` - shows you the differences between your current files and the last commit (what changes you have made)

-   `git push` - pushes your local commits to GitHub

-   `git pull` - pulls the latest changes from GitHub to your local repository

-   `git commit filename -m "your commit message"` - commits your changes to the file with a message describing what you did.

-   `git commit -a -m "your commit message"` - commits all your changes with a message describing what you did.

-   `git add filename` - stages a file for commit (you need to stage files before you can commit them)

-   `git checkout -- filename` - discards changes in a file and reverts it to the last committed version (use with caution, as this will lose any unsaved changes in that file)

-   `git branch branchname` - creates a new branch with the name "branchname"

-  `git checkout branchname` - switches to the branch "branchname"

- `git merge branchname` - merges the branch "branchname" into the current branch


## Assignment: Git and Stream Temperature


For this assignment, we are going to download some USGS data on stream temperature from the Willametter River in Oregon

Here's some R code to download it, 
* X_00010_00003 is the water temperature in C

```{r}
install.packages("dataRetrieval")
library(dataRetrieval)

siteNumbers <- c("14211720", "14211010", "14166000")
parameterCd <- "00010"  # water temperature
startDate <- "2010-01-01"
endDate <- "2023-12-31"

temp_data <- readNWISdv(siteNumbers, parameterCd, startDate, endDate)
colnames(temp_data) <- c("agency_cd", "site_no", "Date", "temp_C")
temp_data$month = month(temp_data$Date, label=TRUE)
temp_data$year = year(temp_data$Date)

# check by plotting
ggplot(temp_data, aes(x=Date, y=temp_C, color=site_no)) +
  geom_line() +
  labs(title="Stream Temperature at Willamette River Sites",
       x="Date", y="Temperature (°C)", color="Site Number") +
  theme_minimal()
```


Working together create a new repository

* add a quarto file, with nice documentation, that downloads the data (call it something informative)
* also store the data in the repository (e.g. as a csv file) so that you can work with it without having to download it again
* add the option to load the data to your quarto file 
* add a code chunk that plots the data (e.g. using ggplot)
* add a code chunk that calculates some summary statistics (e.g. mean, max, min temperature for each site)

Make sure you push to *GitHub*. 

Now imagine you are working on a project, where you are looking trends in water temperature you want to figure out if a) there is a trend and b) if there slope (trend)

Have each partner focus on trends for a particular month

Here's some example code for how you might compute the trend of temperatures in a given month, the slope of the GLM model will tell you what the trend is (how much temperature increase/decreases per year)  - And site_no coefficient is telling you how much the temperature differs between sites (relative to the reference site which is site_no 14166000 in this case)

```{r}
library(tidyverse)
# filter for July temperatures
july_data <- subset(temp_data, month == "Jul")
# fit a GLM model to see if there is a trend in July temperatures over time
july_model <- glm(temp_C ~ year+site_no, data = july_data)
summary(july_model)

```


Each partner creates their own branch, and does their own analysis based on a particular month
(if you want to can expand a bit and compare trends between site)

Then merge your branches back into main and create a summary document (quarto), that
sumamrizes your finding and adds one plot to illustrate what you think might be the most meaningful result for a manager


Turn in your *GitHub* repo link on Canvas  


Grading Rubric

* Code is well documented and organized (20 points)
* Created a new repository and added a quarto file with code to download, load, and analyze the data (20 points)
* Each partner created a branch and analyzed trends for a particular month (20 points)
* Merged branches back into main  (20 points)
* Summary plot and analysis (20 points)








