---
title: 'Data Types and Structures '
format:
  revealjs:
    theme: solarized
    resources: ["images"]
editor: visual
execute:
  echo: true
  warning: true
  message: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo =TRUE)
knitr::opts_chunk$set(error=TRUE)

library(tidyverse)

```

# <span style="color:blue"> Data types

All programming languages use data-types, or structures to hold information

-   integer
-   floating point/ real / numeric
-   character
-   string

Often data types are multi-dimensional Some useful ones in R

-   vector
-   matrix
-   data frame
-   tibble
-   factors
-   lists

Programming often involves selecting and building data structures

Good data structures are

-   as simple as possible
-   easy to understand (readable names)
-   easy to manipulate
-   easy to visualize

## Ways to generate data {.scrollable}

For testing your code, often helpful to generate *test* data

We also often generate random data to explore the behavior of our code and to understand how it works

    -   *runif(min,max)* (sampling from a uniform distribution)
    -   *rnorm(mean,sd)* (sampling from uniform and normal distributions)
    -   *rpois(lambda)* (sampling from a Poisson distribution)
    
(many other options - here's a good *cheatsheet* for random number generation in R 

[Random Generation](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/Randomh)

Other ways to generate data

    -   *sample(vector,size,replace)* (sampling from a vector of values)
    
    -   *seq(start,end,by)* (for creating a sequence of numbers)


## Example

```{r}
# want to random start 20  fires in a fire model, locations numbered from 1 to 100
firestarts = runif(20, min=0, max=100)
ggplot(data.frame(firestarts=firestarts), aes(firestarts))+geom_histogram()
# want to mimic a population of 20 otters with a mean weight of 10 kg and a standard deviation of 2 kg
otterweight = rnorm(20, mean=10, sd=2)
ggplot(data.frame(otterweight), aes(otterweight))+geom_histogram()
# notice how the more you sample the more it looks normal
otterweight = rnorm(2000, mean=10, sd=2)
ggplot(data.frame(otterweight), aes(otterweight))+geom_histogram()


# sample allows you to pick from a vector of values, lets say we want to randomly pick years from a sequence of years from 2000 to 2025
# two options - we allow a year to be picked more than once
pick = sample(seq(2000,2025), size=20, replace=T)
pick
# no repeat years
pick = sample(seq(2000,2025), size=20, replace=F)
pick

```

# Data in R Packages

* Many Packages in R have datasets associated with them

* To see this you can use the **data()** command

* To load a dataset from a package you can use **data(datasetname)**

Lets try and then use *sample* to sample from one

```{r}

data()

# Lets look at New York Air Quality Measurement
# load into your R environment
data(airquality)
# learn about dataset
help(airquality)
# a good way to see structure of data
str(airquality)

# does Ozone correlate with air temperature
ggplot(airquality, aes(x=Temp, y=Ozone))+geom_point()

# lets say we are testing a new product and we want a sample of 5 ozone measurement that conforms to the distribution in this dataset
ozone_sample = sample(airquality$Ozone, size=5, replace=F)
ozone_sample

# plot our sample days

ggplot(airquality, aes(x=Temp, y=Ozone))+geom_point()+
  geom_point(data=data.frame(Temp=airquality$Temp[ozone_sample], Ozone=ozone_sample), color="red", size=3)

```

# <span style="color:blue"> Data Types: Factors \span

something that has different **classes** or **groups** useful for doing calculations with categories

Here's an example:

First lets look at a standard numeric vector

```{r}
a = c(1.3, 1, 4, 1.3, 22)
# compute the mean
mean(a)
```

What if **a** is a factor

What do commands like **mean** do

```{r}
a = as.factor(a)
# compute the mean
mean(a)

#why? lets look
a




```

We can use **summary** with factors to get frequencies in each category (or “level” )

```{r fishes}

# create vector of possible fish 
possible.fish = c("salmon","steelhead","shark","tuna","cod")

# we can use sample to simulate a random recording of catch by fisherman, lets say we pick 20 fish from the net

catch1 = base::sample(possible.fish, size=20, replace=T)
# because possible.fish was a factor catch1 will be a factor
catch1

summary(catch1)
# if we want summary to be more useful - make this a factor
catch1 = as.factor(catch1)


# to quickly get frequencies of different fish and to plot 
summary(catch1)
plot(catch1, col="blue")


# we can also use summary to explore and return information about the distribution
# mean frequency of a particular type of fish
mean(summary(catch1))

# maximum frequency
max(summary(catch1))

# which fish was most frequently caught
which.max(summary(catch1))

#to get just the name 
names(which.max(summary(catch1)))

```

## On your own

In a quarto document \* create a new population of some creature (could be mythical) \* sample from the population \* use summary to explore the distribution of your sample

## Strings and how to work with them

Strings are character data types that are used to represent text

All programming languages have ways to deal with strings
Some commonly used R functions
-   **paste** and **paste0** for concatenating strings together
-   **strsplit** for splitting strings into parts
-  **sprintf** for creating formatted strings 
-   **gsub** for replacing parts of strings
-   **tolower** and **toupper** for changing case
-   **nchar** for counting the number of characters in a string

Most of these are easy to use - just look at help for the function and try it out

**sprintf** however takes some specialized knowledge

## Creating strings with sprintf

**sprintf** used something called
*format string syntax* to create formatted strings

*format string syntax* is used in many places -for searching for and creating strings - AND - its used by many programming languages 

- C / C++ → printf
- Python → old-style % formatting and str.format()
- Java / JavaScript → similar formatting ideas
- SQL (some dialects), 
- Unix Shell (terminal window) 

## format string syntax

-   **%s** replace with a string
-   **%d** replace with an integer value
-   **%f** replace with a real value
-   **%4.1f** replace with a real value with 4 digits, two after decimal
-   **\\n** add a line return


## Example

make a string with **sprintf** and add to plot title


```{r}
# use results for creating text
# sprintf creates a string %s mean use what ever is after the , to fill in a string
plottitle=sprintf("We like %s", names(which.max(summary(catch1))))

ggplot(data.frame(catch1=catch1), aes(catch1, fill=catch1))+geom_histogram(stat="count")+labs(title=plottitle)

# you can also add numbers to the string
plottitle=sprintf("We mostly caught %s \n max catch(%d)", names(which.max(summary(catch1))), max(summary(catch1)))
ggplot(data.frame(catch1=catch1), aes(catch1, fill=catch1))+geom_histogram(stat="count")+labs(title=plottitle)

```

## But I don't want a factor

Sometimes you don't want factors and R thinks something should be How to change back? **as.numeric** makes sense ...but

```{r, echo=TRUE}

a = as.factor(c(1.3, 1, 4, 1.3, 22))
#sum(a)

# try to make a numeric version from the factor
b = as.numeric(a)
sum(b)
b

# better
b = as.character(a)
b = as.numeric(b)
b
sum(b)
```

# <span style="color:blue"> Lists \span


-   Lists are the most “informal” data structures in R
-   List are really useful for keeping track of and organizing groups of things that are not all the same
-   A list could be a table where number of rows is different for each column
-   A list can have numeric, character, factors all mixed together
-   List are often used for returning more complex information from function (e.g. lm)

```{r introlist, echo=TRUE}

# make a list
sale = list(id=2, quality="high", contents=c("apple","cherry"), cost=c(4,5))
sale

#ways to access elements
sale$id
sale$what

# you can also access by list item number
# the [x] denotes the xth item in the list
sale[[3]]
sale[["contents"]]


# how do you get the second element in the vector that has the contents
# there are two ways


# add to a list
sale$location = "Farmers market"
sale
# or remove
sale$location = NULL
sale

# some tricky things
# correct accessing items in list
sale$cost
sale[[4]]

# works but
#sale[4]


sum(sale$cost)
sum(sale[[4]])

```

# What you've learned

-   common data types
-   ways to generate data 
-   string format syntax
-   lists

# <span style="color:orange">PACKAGES from Git/span\>

-   **Package**: An extension of the R base system with functions, data and documentation in standardized format.
-   **Library**: A directory containing installed packages.
-   **Repository**: A website providing packages for installation.
-   **Source**: The original version of a package with human-readable text and code.
-   **Binary**: A compiled version of a package with computer-readable text and code, may work only on a specic platform.
-   **Base packages**: Part of the R source tree, maintained by R Core.
-   **Recommended packages**: Part of every R installation, but not necessarily maintained by R Core.
-   **Contributed packages**: All the rest. This does not mean that these packages are necessarily of lesser quality than the above, e.g., many contributed packages on CRAN are written and maintained by R Core members. They simply try to keep the base distribution as lean as possible.
-   **User packages**: Packages that you write, share with a smaller community - not downloadable from CRAN, but can be downloaded (or sent as a zip file) and loaded into R

## [Installing packages from GitHub]{style="color:orange"}

you can install user packages from *github* - you need the *devtools* library to do this

install_github("naomitague/rainbow")

*Important* this doesn't create a *repo* of rainbow inside you working repository - so its safe! But you cannot *clone* or create a *repo* inside the directory structure of another one
